---
title: "431 Class 04"
author: Thomas E. Love, Ph.D.
date: "2026-09-03"
format:
  revealjs: 
    theme: dracula
    embed-resources: true
    self-contained: true
    slide-number: true
    footnotes-hover: true
    preview-links: auto
    date-format: iso
    footer: "431 Class 04 | 2026-09-03 | <https://thomaselove.github.io/431-2026/>"
---

## Today's Agenda

Today, we'll work with some data from the 15-item survey from Class 2.

- Ingesting data from a .csv file
- Data management and clean-up
- Dealing with missing data via "complete cases"
- Addressing a few exploratory comparisons
- Visualization, Modeling and Inference

::: aside

One goal is to make sure you're ready for Lab 1, due 2026-09-09.

:::

## How did ingesting the data go?

The 2026 results ([pdf of survey is here](https://github.com/THOMASELOVE/431-classes-2026/blob/main/class02/class02-survey-431-2026.pdf)):

- I received 41 paper surveys.
- Two surveys had identical responses, so I dropped one.

In all, 14 of the 41 (**34.1%**) responses gave me no trouble in 2026. In 2014-2025, 36% of forms gave me no trouble.

::: aside

Issues seen in 2026: interpreting handwriting, imprecise circles, cross-outs, not following directions on pulse rate, problems interpreting the handedness scale, difficult to identify units of measurement, listing multiple colors

:::

## What have I done to set up today?

From 431-data, I have:

- `431-first-r-template.qmd`: our Quarto starting point
- `Love-431.R` script (for the `lovedist()` function)
- `class02_survey_2014_to_2026.csv`: main data file

Then I created an R Project directory for this work, which has the Quarto template in the main directory, and the .R and .csv files in a "c04/data" subdirectory.


## Next steps

1. Open RStudio
2. Start a new R Project in the Project Directory I created on the previous slide.
3. Open the Quarto template, then saved it as `class04_survey_complete.qmd` after adjusting the title and author^[This file is also available to you on our [431-data page](https://github.com/THOMASELOVE/431-data).].
4. Next step is to adjust the R setup and get rolling.

## Load packages and set theme

```{r}
#| echo: true
#| message: false

knitr::opts_chunk$set(comment = NA)

source("c04/data/Love-431.R") 

library(janitor)
library(naniar)    ## this will help us identify missing values
library(patchwork) ## stitch together two different ggplots
library(easystats)
library(tidyverse)

theme_set(theme_bw())
```

## Ingest 2014-2026 data

```{r}
#| echo: true

s15 <- read_csv("c04/data/class02_survey_2014_to_2026.csv",
                show_col_types = FALSE) |>
  janitor::clean_names()

s15
```

## Variables in the data

```{r}
#| echo: true
names(s15)
```

Let's **select** just the variables we'll use today.

```{r}
#| echo: true
s15 <- s15 |> select(student, glasses, english, love_htcm, 
                     smoke, haircut, height_in, hand_span, 
                     favcolor, lastsleep, pulse, year, 
                     colors_13) 

dim(s15)
```

- Same number of rows, but now only 13 columns.

## What is in these variables? (1/2)

Variable | Description
---------: | :------------------------------------------------
`student` | identifying code (202601, 202602, etc.)
`glasses` | *y* if student wears corrective eyewear, otherwise *n*
`english` | *y* if english is best language, otherwise *n*
`love_htcm` | guess of my height (in centimeters)
`smoke` | *1* = never smoked, *2* = quit, *3* = current smoker
`haircut` | price (in $) of last haircut

## What is in these variables? (2/2)

Variable | Description
---------: | :------------------------------------------------
`height_in` | your height (in inches)
`hand_span` | your hand span (in centimeters)
`favcolor` | your favorite color (>40 colors mentioned)
`lastsleep` | hours of sleep last night
`pulse` | current pulse rate (beats per minute)
`year` | year of survey (2014 - 2026)
`colors_13` | my attempt to collapse the colors a bit

## Variable types, as ingested

- `<chr>` = character variable: as opposed to a *factor*
- `<dbl>` = double-precision numeric variable

```{r}
#| echo: true
glimpse(s15)
```


## The subject ID should be of the character type

Some of the variables need to be of a different type:

- `student` is just a code to identify each subject. We'd like that to be of `character` type, rather than `double`, so we don't accidentally use it in a model.

```{r}
#| echo: true

s15 <- s15 |> mutate(student = as.character(student))
```

## Making some variables into factors

- `glasses`, `english`, `favcolor`, and `colors13` should all be *factors*, rather than character variables.
    - R made these characters because the values included non-numbers (specifically, letters.)
    - A *factor* is easy to use in an analysis or summary, while a *character* variable isn't.

```{r}
#| echo: true

s15 <- s15 |> mutate(glasses = factor(glasses),
                     english = factor(english),
                     favcolor = factor(favcolor),
                     colors_13 = factor(colors_13))
```

## `smoke` is also categorical

We'd also like to turn `smoke` into a factor, and use the actual names for the levels rather than the 1-2-3 codes.

```{r}
#| echo: true
s15 <- s15 |> mutate(smoke = fct_recode(factor(smoke), 
                                        "Never" = "1", 
                                        "Quit" = "2", 
                                        "Current" = "3"))

s15 |> tabyl(smoke) |> adorn_pct_formatting()
```

## Missing values?

- The `read_csv()` function recognizes (by default) NA or blank cells as missing. How many cells are missing?

```{r}
#| echo: true

dim(s15)
696 * 13
n_miss(s15)
pct_miss(s15)
```

We have 869 missing cells in our rectangular tibble, so 9.6% of the data are missing.

## Missingness by variable?

::: aside

- I started asking for guesses of my height in *2021*.
- I started asking about glasses in *2019*.

:::

```{r}
#| echo: true
miss_var_summary(s15)
```

## Missingness by row (case / subject)?

```{r}
#| echo: true
miss_case_table(s15)
```

::: aside

- About 42% of the cases are complete.

:::

## Quick Summary of All Variables

```{r}
#| echo: true

summary(s15)
```

## Describing Quantities

```{r}
#| echo: true

describe_distribution(s15)
```


## Exploratory Question 1

What is the relationship between **hand span** and **height** for 431 students since 2014? Can we draw a scatterplot?

```{r}
#| echo: true
#| warning: true
ggplot(data = s15, aes(x = height_in, y = hand_span)) +
  geom_point()
```

## Filtering to complete cases

Let's require complete data on these variables...

```{r}
#| echo: true

s15_q1 <- s15 |> filter(complete.cases(hand_span, height_in))

s15_q1 |> reframe(lovedist(hand_span))
s15_q1 |> reframe(lovedist(height_in))
```

## Scatterplot of Hand Span by Height

```{r}
#| echo: true

ggplot(data = s15_q1, aes(x = height_in, y = hand_span)) +
  geom_point()
```

## Lots of ties - can we show that?

```{r}
#| echo: true

ggplot(data = s15_q1, aes(x = height_in, y = hand_span)) +
  geom_jitter()
```

## Add a loess smooth, improve labels

```{r}
#| echo: true
#| output-location: slide

ggplot(data = s15_q1, aes(x = height_in, y = hand_span)) +
  geom_smooth(method = "loess", col = "blue", formula = y ~ x, se = TRUE) +
  geom_point() +
  labs(x = "Height (in)", y = "Hand Span (cm)", 
       title = "Association of student hand span and height",
       subtitle = "n = 618 students in 431 between 2014 and 2026")
```

## Add a linear model, as well

```{r}
#| echo: true
#| output-location: slide

ggplot(data = s15_q1, aes(x = height_in, y = hand_span)) +
  geom_smooth(method = "loess", col = "blue", 
              formula = y ~ x, se = FALSE) +
  geom_smooth(method = "lm", col = "red", 
              formula = y ~ x, se = TRUE) +
  geom_point() +
  labs(x = "Height (in)", 
       y = "Hand Span (cm)", 
       title = "Association of student hand span and height",
       subtitle = "n = 618 students in 431 between 2014 and 2026")
```

## What is the straight line fit?

```{r}
#| echo: true

fit1 <- lm(hand_span ~ height_in, data = s15_q1)

model_parameters(fit1, pretty_names = FALSE, ci = 0.95) 
```

So what is the straight line equation from `lm`?

:::{.incremental}

- **hand_span** = -0.87 + 0.32 **height_in**
- How do we interpret the slope (0.32) of `height_in`?
- How do we interpret the intercept (-0.87)?

:::

::: aside

Note the 95% confidence intervals around these OLS estimates.

:::

## Exploratory Question 2

What is the relationship between **haircut** price and **height** among the 40 students responding in 2026?

- What do we expect to see in this scatterplot? Why?

Let's just look at 2026 data this time.

```{r}
#| echo: true
s15_2026 <- s15 |> filter(year == 2026)
```

- What do the results below tell us?

```{r}
#| echo: true
dim(s15_2026)
s15_2026 |> select(haircut, height_in) |> n_miss()
```

## Haircut Price vs. Height in 2026

```{r}
#| echo: true
#| output-location: slide

ggplot(s15_2026, aes(x = height_in, y = haircut)) +
  geom_jitter() +
  geom_smooth(method = "loess", formula = y ~ x, 
              se = FALSE, col = "blue") +
  geom_smooth(method = "lm", formula = y ~ x, 
              se = TRUE, col = "red") +
  labs(title = "Association of haircut price and height",
       subtitle = "n = 40 students in 431 for Fall 2026 (values jittered)",
       x = "Height (in)", 
       y = "Price of last haircut ($)")
```

## What is the OLS model here?

```{r}
#| echo: true

fit2 <- lm(haircut ~ height_in, data = s15_2026)
model_parameters(fit2, pretty_names = TRUE, ci = 0.90)
```

- How do we interpret the slope of `height_in`?
- How do we interpret the intercept?
- What does `pretty_names = TRUE, ci = 0.90` do?

Are we likely to see a similar result in previous years?

## Compare 2026 to 2014-2025

```{r}
#| echo: true
#| output-location: slide

plot1 <- ggplot(s15_2026, aes(x = height_in, y = haircut)) +
  geom_point() +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE, col = "blue") +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, col = "red") +
  ylim(0, 300) + ## make limits of y axis 0 and 300
  labs(title = "2026: Association of haircut price and height",
       subtitle = "n = 40 students in 431 for Fall 2026", 
       x = "Height (in)",  y = "Price of last haircut ($)")

s15_q2 <- s15 |> filter(complete.cases(height_in, haircut)) |>
  filter(year < 2026)  ## this contains 642 rows, it turns out

plot2 <- ggplot(s15_q2, aes(x = height_in, y = haircut)) +
  geom_point() +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE, col = "blue") +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, col = "red") +
  ylim(0, 300) + ## make limits of y axis 0 and 300 here, too
  labs(title = "2014-2025: Association of haircut price and height",
       subtitle = "n = 642 students in 431 for 2014-2025", 
       x = "Height (in)",  y = "Price of last haircut ($)")

plot1 + plot2 +                 ## "stitch" the plots together
  plot_layout(axes = "collect") ## thanks to patchwork package
```

# Seems like the place for a break.{autoslide=130000}

## Exploratory Question 3

Students specified their height (`height_in`) and guessed mine (`love_htcm`), since 2021.

- Did taller students guess that I was taller?
- Let's convert the heights so each is in inches.

```{r}
#| echo: true
s15_q3 <- s15 |> 
  select(student, height_in, love_htcm) |>
  filter(complete.cases(height_in, love_htcm)) |>
  mutate(love_htin = round_half_up(love_htcm * 0.3937, 1)) |>
  mutate(delta = love_htin - height_in)
```

So what's in this new tibble?

## Contents of `s15_q3` tibble

```{r}
#| echo: true
s15_q3
```

- What does `delta` mean?

## Numerical Summaries

- Student Guesses of My Height (in inches)

```{r}
#| echo: true
s15_q3 |> reframe(lovedist(love_htin))
```

- Student Heights (in inches)

```{r}
#| echo: true
s15_q3 |> reframe(lovedist(height_in))
```

Suppose we want to compare these means.

## Visualizing the Distributions

```{r}
#| echo: true
#| output-location: slide

plot3 <- ggplot(s15_q3, aes(x = love_htin)) +
  geom_histogram(binwidth = 1, fill = "tomato", col = "white") +
  labs(title = "Histogram: Guess of Dr. Love's height (inches)",
       x = "Guess of Dr. Love's height (in)", y = "# of students")

plot4 <- ggplot(s15_q3, aes(x = height_in)) +
  geom_histogram(binwidth = 1, fill = "royalblue", col = "white") +
  labs(title = "Histogram: Student height (inches)",
       x = "Student's height (in)", y = "# of students")

plot3 / plot4 + ## thanks, patchwork!
  plot_annotation(title = "306 students in Dr. Love's 431 Class, 2021-2026") 
```

## Numerical Summaries

Each subject gives us their guess of my height, and their height.

- $\Delta$ = How much taller did I seem? (in inches)

```{r}
#| echo: true
s15_q3 |> reframe(lovedist(delta))
```

These are the "paired" differences. What do they look like?

## Histogram of $\Delta$

```{r}
#| echo: true
#| output-location: slide

ggplot(s15_q3, aes(x = delta)) +
  geom_histogram(binwidth = 1, fill = "aquamarine", col = "black") +
  labs(title = "How much taller did Dr. Love seem?",
       subtitle = "306 students in his 2021-2026 431 classes",
       x = "Guess of Dr. Love's height - Student's height (inches)",
       y = "# of students")
```

## Estimating the mean $\Delta$

Use a linear model to estimate the true mean difference between guesses of my height and actual student heights, using a **90%** confidence interval.

```{r}
#| echo: true

fit3 <- lm(delta ~ 1, data = s15_q3)
model_parameters(fit3, ci = 0.90)
```

- What would a **95%** confidence interval look like?

```{r}
model_parameters(fit3, ci = 0.95)
```

## Did taller people guess I'm taller?

```{r}
#| echo: true

ggplot(s15_q3, aes(x = height_in, y = love_htin)) +
  geom_jitter() +
  geom_smooth(method = "loess", formula = y ~ x, col = "blue")
```


## Exploratory Question 4

Did students for whom English is their primary language have larger (or smaller) hand spans?

```{r}
#| echo: true

s15_q4 <- s15 |> filter(complete.cases(english, hand_span)) |>
  select(student, english, hand_span, year)

dim(s15_q4)
summary(s15_q4)
```

## Visualizing the groups

```{r}
#| echo: true
#| output-location: slide

ggplot(s15_q4, aes(x = hand_span, y = english)) +
  geom_violin(aes(fill = english)) +
  geom_boxplot(width = 0.4) +
  stat_summary(fun = "mean", geom = "point", 
               size = 3, col = "royalblue") +
  labs(title = "Comparing hand span by preferred language",
       subtitle = "613 of Dr. Love's 431 students asked 'Is English your most comfortable language?'",
       caption = "Sample means indicated by blue dots",
       x = "Hand Span (in cm)",
       y = "Prefers English?")
```

## Summarizing our Samples

- Compare the English "Yes" vs. "No" group on **hand span**.

```{r}
#| echo: true

s15_q4 |> group_by(english) |>
  reframe(lovedist(hand_span))
```

- What seems to be happening here?
- Does the difference in sample means look important?

## Build a model

Let's use a linear model to estimate the difference in hand span between those whose primary language is English and those whose primary language is not English...

```{r}
#| echo: true
fit4 <- lm(hand_span ~ english, data = s15) 

model_parameters(fit4, ci = 0.90)
```

- How would you describe the point estimate here?
- What is the 90% confidence interval?

## Some Categories: Your Favorite Colors {.smaller}

```{r}
#| echo: true
s15 |> tabyl(favcolor) |> arrange(desc(n))
```

## I collapsed these to 13 colors

```{r}
#| echo: true
s15 |> tabyl(colors_13) |> arrange(desc(n)) |> 
  adorn_totals() |> adorn_pct_formatting()
```

## Glasses vs. English

This is a 2x2 table.

```{r}
#| echo: true
s15 |> tabyl(glasses, english)

s15 |> 
  filter(complete.cases(glasses, english)) |>
  tabyl(glasses, english) |>
  adorn_totals(where = c("row", "col"))
```

## Smoking Status by English

Here's a 3x2 table.

```{r}
#| echo: true
s15 |> filter(complete.cases(smoke, english)) |>
  tabyl(smoke, english) |>
  adorn_totals(where = c("row", "col")) |> knitr::kable()
```

## Finishing Up

- I've shown you everything you need to complete [Lab 1](https://github.com/THOMASELOVE/431-labs-2026/tree/main/lab1), due Wednesday 2026-09-09 at noon.
- `class04_survey_complete.qmd`: Quarto file with the analyses in these slides, pulling directly from [our 431-data page](https://github.com/THOMASELOVE/431-data). You can render to HTML if you like.
- As always, you have the [Quarto file for these slides](https://github.com/THOMASELOVE/431-classes-2026/tree/main/class04#todays-slides), too.

**Next week** we'll work with new data and get into chapters 5-7 of [my book](https://thomaselove.github.io/431-book/).

## Session Information

```{r}
#| echo: true
xfun::session_info()
```


