---
title: "431 Class 15"
author: Thomas E. Love, Ph.D.
date: "2026-10-13"
format:
  revealjs: 
    theme: dracula
    embed-resources: true
    self-contained: true
    slide-number: true
    footnotes-hover: true
    preview-links: auto
    date-format: iso
    footer: "431 Class 15 | 2026-10-13 | <https://thomaselove.github.io/431-2026/>"
---

## Today's Agenda

- Power and Sample Size, Part 2 (originally in Slide set 14)
- Some Project A Advice
- Your Questions about Project A and Quiz 1

### Today's R Package

Just one package today.

```{r}
#| echo: true
#| message: false

knitr::opts_chunk$set(comment = NA)

library(pwr)
```

# Power and Sample Size: Comparing Proportions

## Designing a New Study

> (PI): In our pilot, we saw $p_{groupA}$ = 0.18 and $p_{groupB}$ = 0.24. Help me design a new study using a two-sided test with $\alpha = 0.05$. This time, let's have as many in groupA as in groupB. We should have 90% power to detect a difference almost as large as what we saw in the pilot, or larger, so a difference of 6 percentage points. 

What sample size is required to achieve these aims?

## How `power.prop.test` works

We specify 4 of the following 5 elements of the comparison, and R calculates the fifth.

-   The sample size (interpreted as the \# in each group, so half the total sample size)
-   The true probability in group 1
-   The true probability in group 2
-   The significance level ($\alpha$)
-   The power (1 - $\beta$)

Requires you to work with balanced designs.

## Using `power.prop.test`

To find the sample size for a two-sample comparison of proportions using a balanced design:

-   use a two-sided test, with $\alpha$ = .05, and power = .90,
-   we estimate that groupA has probability .18 of our outcome,
-   and we will try to detect a difference between this group and groupB, who we estimate will have a probability of .24 of our outcome.

Any guess as to needed sample size?

## Finding the required sample size

```{r}
#| echo: true
#| eval: false
power.prop.test(p1 = .18, p2=.24, alternative = "two.sided",
                sig.level = 0.05, power = 0.90)
```

## Answer

```{r}
#| echo: true
power.prop.test(p1 = .18, p2=.24, alternative = "two.sided",
                sig.level = 0.05, power = 0.90)
```

- We'd need at least 967 groupA subjects, and 967 more in groupB to accomplish these goals: a total of 1934 subjects.

## A One-Sided Test?

Suppose we get 400 groupA and 400 groupB subjects. With what power could we detect the difference between 0.18 in one group and 0.24 in the other, in a *one-sided* $\alpha$ = .10 test?

```{r}
#| echo: true
#| eval: false
power.prop.test(n=400, p1=.18, p2=.24, sig.level = 0.10,
                alternative="one.sided")
```

## Answer

Suppose we get 400 groupA and 400 groupB subjects. With what power could we detect the difference between 0.18 in one group and 0.24 in the other, in a *one-sided* $\alpha$ = .10 test?

```{r}
#| echo: true
power.prop.test(n=400, p1=.18, p2=.24, sig.level = 0.10,
                alternative="one.sided")
```


## Using the `pwr` package to assess sample size for Unbalanced Designs

The `pwr.2p2n.test` function in the `pwr` package can help assess the power of a test to determine a particular effect size using an unbalanced design, where $n_1$ is not equal to $n_2$.

As before, we specify four of the following five elements of the comparison, and R calculates the fifth.

## Now the five elements are...

-   `n1` = The sample size in group 1
-   `n2` = The sample size in group 2
-   `sig.level` = The significance level ($\alpha$)
-   `power` = The power (1 - $\beta$)
-   `h` = the effect size, which can be calculated separately in R based on the two proportions being compared: $p_1$ and $p_2$.

## Calculating the Effect Size `h`

To calculate the effect size for a given set of proportions, use `ES.h(p1, p2)` which is available in the `pwr` package.

For instance, comparing .18 to .24, we have the following effect size.

```{r}
#| echo: true
ES.h(p1 = .18, p2 = .24)
```

Suppose we can recruit 800 subjects in groupA but only 400 in groupB.

## What power would we have?

Detect the distinction between p1 = .18 and p2 = .24 with a 90% confidence level (two-sided) with 800 subjects in groupA and 400 in groupB?

```{r}
#| echo: true
pwr.2p2n.test(h = ES.h(p1 = .18, p2 = .24), n1 = 800, n2 = 400, 
                   sig.level = 0.1)
```

## Comparison to Balanced Design

How does this compare to the results with a balanced design using 1200 subjects in total, i.e. 600 per group?

```{r}
#| echo: true
pwr.2p2n.test(h = ES.h(p1 = .18, p2 = .24), n1 = 600, n2 = 600, 
                   sig.level = 0.10)
```

## We could instead have used...

```{r}
#| echo: true
power.prop.test(p1 = .18, p2 = .24, sig.level = 0.10, n = 600)
```

Each approach we've used includes approximations, and slightly different ones, so it's not surprising that the answers are similar, but not identical.

## Some people will drop out (1)

Suppose my calculation says I need 1000 subjects in my study. How many subjects do I need to recruit?

- Suppose I expect at most 20% of recruits to drop out before the study ends.
- If I expect at most 20% of recruits to drop out, then perhaps I should recruit enough that if 20% leave, I still have 1000.

but ...

## Some people will drop out (2)

Suppose my calculation says I need 1000 subjects in my study. How many subjects do I need to recruit?

- If I expect at most 20% of recruits to drop out, that means I would retain at least 80% of the subjects.
- So I recruit R = 1000 / 0.8 = 1250 people. If I retain 80% of my 1250 recruits, that's 0.8 (1250) = 1000 subjects.

## Power Calculations

Most common scenario: identify desired significance level $\alpha$ and desired power (1 - $\beta$) and the details of the plan in terms of the comparison to be made, and how the data are to be collected.  

Then calculate a minimum necessary sample size to achieve these desires, with $\alpha = 0.05$ and $\beta = 0.2$ commonly used.

- Neither 95% confidence nor 80% power is a magical choice.
- Anything below 80% power is hard to justify in real work.

## A useful metaphor?

Sometimes I like to think of science as a march towards a destination. 

Actually, I suppose it's an infinitely long march towards an ever-receding destination, but let's leave the philosophy out of it for a moment. 

Suppose, for example, that we're trying to make a meaningful change in the world, perhaps to treat an infection. 

What we're trying to do is related to where we are in the March of Science.

## Early in the March of Science

In **early** work, we're focused more on discovery than making final decisions.

- We don't have a lot of past experience, so we bring little relevant data to the table. 
- We're (often) most concerned about discovering new possibilities, and we don't have a very clear sense of where to go next.
- We're (often) less concerned about false starts than we are about missed opportunities.

## Late in the March of Science

In **late** work, we're focused more on making a decision about how to treat.

- We have a fair amount of relevant history to draw on, sometimes quite detailed.
- We're more concerned about testing the limits of our current knowledge than we are about missing opportunities to consider a new pathway.
- We're often concerned about doing harm if we implement the strategy that looks most promising.
  
## Power and "significance"?

If we treat our sample size and study design as fixed strategies, then we must choose between:

- reducing $\alpha$, the rate of Type I error (increasing our confidence) and
- reducing $\beta$, the rate of Type II error (increasing our power)

## A Little Example

Suppose we are testing a new treatment for some condition.

- A Type I error means we conclude this treatment is helpful, when it actually isn't.
- A Type II error means we conclude this treatment is not helpful, when it actually is.

## Early Work: Power & Sample Size (1)

In early work, we are searching for treatments of promise, and our initial study will inevitably not be the last word on the subject, but rather will be followed up by confirmatory studies. In such a setting, it is often the case that:

- We're not so concerned about getting results that cause us to continue to explore a treatment that doesn't actually do what we need it to do.
- We're really concerned about ruling out a treatment that is promising before we should. 

## Early Work: Power & Sample Size (2)

This implies we should prioritize reducing Type II error rates (we want more power to detect small but real effects, even if this means we will occasionally identify something as promising when it isn't.)

This means setting lower confidence levels and higher power levels, potentially, than the standard 95% confidence and 80% power.

## Late Work: Power & Sample Size (1)

In late work, we have already identified promising treatments, and we are trying to confirm those results. The current study may actually be the last word on the subject, and we want to be sure we do no harm.

- We're very concerned about getting results that cause us to continue to explore a treatment that doesn't actually do what we need it to do.
- We're less concerned about ruling out a treatment that is promising but doesn't actually work. 

## Late Work: Power & Sample Size (2)

This implies we should prioritize reducing Type I error rates (we want greater confidence, even at the expense of power, that the effect we claim based on past data holds up.) 

This kind of confirmatory work is usually well suited to studies set up with higher confidence (perhaps 95% or 99% or more) and lower power (80% is the minimum I would recommend) against reasonable alternatives. 

## Conclusions 1-2

1. If you're early in the March of Science (perhaps just one pilot study has been done) then I would emphasize Type II error (power) more than usual, perhaps pushing required power to 90%, at least for a reasonably substantial "minimum scientifically important difference" $\delta$.

2. If you're late in the March of Science (perhaps confirming the results of multiple prior studies) then I would be happier with 80% power and higher levels of confidence.

## Conclusions 3-4

3. If you're in the middle, trading off Type I and Type II error is worth some thought, but I'd never recommend being under 80% power for an effect that matters.

4. If it's feasible to run a study large enough to have strong performance on both $\alpha$ and $\beta$, that's obviously ideal. Typically that doesn't happen in early work.

## So ... How Big A Sample Do I Need?

1. What is your research question?
2. What is the budget?

# Standing Break

# Project A

## Project A due 2026-10-21 at noon

1. Your project A master data set (with missing values), saved as an .Rds file.
2. A detailed report, written in Quarto and rendered in HTML (17 sections), which describes all of your work and follows [the Quarto template provided](https://raw.githubusercontent.com/THOMASELOVE/431-data/refs/heads/main/data/project_A/431-projA-template-2026.qmd).
3. A 3-minute recorded video presentation which highlights some key findings from your report ([details here](https://thomaselove.github.io/431-projectA-2026/#the-highlight-video)), and
4. A [self-evaluation form](https://thomaselove.github.io/431-projectA-2026/#the-self-evaluation), which you’ll complete after submitting the other three elements to Canvas.

## Project A Advice (1)

The most important parts of the Report are:

- accurate (and effectively presented) data development, 
- accurate (and effectively presented) analyses, and 
- the appropriateness of your research questions and 
- your responses to them in light of your analyses for each of the four comparisons in Sections 10-13. 
- The reflection in Section 14 is also important.

## Project A Advice (2)

All videos should include a clear statement of the research questions for each of the two analyses presented, and justify the responses to those questions with results from the analyses. 

- Many of the comments Professor Love builds in response to your Project A will relate to the Video. 
  - Do not treat it lightly. It is a vital part of the work, despite its short length.
- Detailed instructions for the Video [are found here](https://thomaselove.github.io/431-projectA-2026/#the-highlight-video). Follow them closely.

## Project A Advice (3)

Perform the elements of the data science cycle (ingestion, tidying, transforming, visualizing, modeling and communicating) effectively. 

- Presenting irrelevant work that does not address the issues in question will hurt your grade.
- Also, rehearse your video and make sure you present clearly and understandably throughout. 
- Make clear choices about what to include - that’s key.
- Do not just show us everything you did.

## Project A Advice (4)

A stronger grade will be associated with doing the tasks we’ve discussed more effectively. 

- Adding a lot of additional and unnecessary analyses/work without editing down to the important issues will not improve your grade, but rather have the opposite effect. 
- We want to see polished, accurate, clear work. 
- Making things clearer is the hardest thing to do, and so it’s the thing that impresses us most.

# Your Questions (about Project A or Quiz 1)?

## Session Information

```{r}
#| echo: true
xfun::session_info()
```
