---
title: "431 Class 07"
author: Thomas E. Love, Ph.D.
date: "2026-09-15"
format:
  revealjs: 
    theme: dracula
    embed-resources: true
    self-contained: true
    slide-number: true
    footnotes-hover: true
    preview-links: auto
    date-format: iso
    footer: "431 Class 07 | 2026-09-15 | <https://thomaselove.github.io/431-2026/>"
---

## Today's Agenda

- What does "Normalish" data look like?
- Comparing Two Samples
  - Paired vs. Independent Samples
  - Checking Assumptions
  - How should we interpret an uncertainty interval?
- Transforming an Outcome (see [Chapter 7](https://thomaselove.github.io/431-book/))
  - The importance of power transformations
  - Tukey's ladder and the Box-Cox plot

## Load packages and set theme

```{r}
#| echo: true
#| message: false

knitr::opts_chunk$set(comment = NA)
source("c07/data/Love-431.R") # for the lovedist() function

library(janitor)
library(patchwork)
library(readxl)
library(knitr)
library(kableExtra)
library(MKinfer)
library(rstanarm)
library(car)         
library(easystats)
library(tidyverse)

theme_set(theme_bw())
```

## A Simulated Data Set

```{r}
#| echo: false

## Simulation Code

set.seed(431)
dat07 <- tibble(
  var1 = rnorm(n = 125, mean = 100, sd = 15) + 4,
  var2 = 100 + 7.5*rt(n = 125, df = 2, ncp = 0),
  var3 = 0.5*rt(n = 125, df = 5, ncp = 90) + 48,
  var4 = runif(n = 125, min = 80, max = 120) + 3,
  var5 = 17+100*rbeta(n = 125, shape1 = 5, shape2 = 1),
  var6 = rnorm(n = 125, mean = 96, sd = 18) + 6,
)
dat07$var1[88] <- 54.41
dat07$var1[47] <- 149.32
dat07$var2[123] <- 99

## end of Simulation Code
```

```{r}
#| echo: true
dat07
```


::: aside

Simulation details for `dat07` tibble are in the Quarto code.

:::

## Numerical Summaries

```{r}
#| echo: true

res1 <- dat07 |> reframe(lovedist(var1)) |> mutate(variable = "var1")
res2 <- dat07 |> reframe(lovedist(var2)) |> mutate(variable = "var2")
res3 <- dat07 |> reframe(lovedist(var3)) |> mutate(variable = "var3")
res4 <- dat07 |> reframe(lovedist(var4)) |> mutate(variable = "var4")
res5 <- dat07 |> reframe(lovedist(var5)) |> mutate(variable = "var5")
res6 <- dat07 |> reframe(lovedist(var6)) |> mutate(variable = "var6")

rbind(res1, res2, res3, res4, res5, res6) |> relocate(variable) |>
  kable(digits = c(0, 0, 0, 2, 2, 2, 2, 1, 1, 1, 1)) |>
  kable_styling(font_size = 24)
```



## Histograms of `dat07` variables

```{r}
#| echo: true
#| output-location: slide

H1 <- ggplot(dat07, aes(x = var1)) + labs(title = "Histogram of var1") +
  geom_histogram(binwidth = 6, col = "black", fill = "skyblue") 
H2 <- ggplot(dat07, aes(x = var2)) + labs(title = "Histogram of var2") +
  geom_histogram(binwidth = 6, col = "black", fill = "plum")
H3 <- ggplot(dat07, aes(x = var3)) + labs(title = "Histogram of var3") +
  geom_histogram(binwidth = 6, col = "black", fill = "coral") 
H4 <- ggplot(dat07, aes(x = var4)) + labs(title = "Histogram of var4") +
  geom_histogram(binwidth = 3, col = "black", fill = "aquamarine")
H5 <- ggplot(dat07, aes(x = var5)) + labs(title = "Histogram of var5") +
  geom_histogram(binwidth = 4, col = "black", fill = "slateblue")
H6 <- ggplot(dat07, aes(x = var6)) + labs(title = "Histogram of var6") +
  geom_histogram(binwidth = 6, col = "black", fill = "burlywood") 

(H1 + H2 + H3) / (H4 + H5 + H6)
```

## Three-Plot Summary for `var01`

```{r}
#| echo: true
#| output-location: slide

plot1A <- ggplot(dat07, aes(sample = var1)) +
  geom_qq(shape = 21, fill = "skyblue") + geom_qq_line(col = "red") + 
  labs(title = "Normal Q-Q plot of var1", y = "var1", x = "Standard Normal")
plot1B <- ggplot(dat07, aes(x = var1)) +
  geom_histogram(binwidth = 6, col = "black", fill = "skyblue") + 
  stat_function(fun = function(x)
    dnorm(x, mean = mean(dat07$var1, na.rm = TRUE),
          sd = sd(dat07$var1,na.rm = TRUE)) * 
      length(dat07$var1) * 6,  
    geom = "area", alpha = 0.5, fill = "linen", col = "black") +
  labs(title = "Histogram of var1", y = "")
plot1C <- ggplot(dat07, aes(x = var1, y = "")) +
  geom_violin(fill = "skyblue") + geom_boxplot(width = 0.3) +
  stat_summary(fun = "mean", geom = "point", col = "red") +
  labs(title = "Boxplot of var1", y = "", x = "var1")

plot1A + (plot1B / plot1C + plot_layout(heights = c(2,1))) 
```

## Three-Plot Summary for `var02`

```{r}
#| echo: true
#| output-location: slide

plot2A <- ggplot(dat07, aes(sample = var2)) +
  geom_qq(shape = 21, fill = "plum") + geom_qq_line(col = "red") + 
  labs(title = "Normal Q-Q plot of var2", y = "var2", x = "Standard Normal")
plot2B <- ggplot(dat07, aes(x = var2)) +
  geom_histogram(binwidth = 6, col = "black", fill = "plum") + 
  stat_function(fun = function(x)
    dnorm(x, mean = mean(dat07$var2, na.rm = TRUE),
          sd = sd(dat07$var2,na.rm = TRUE)) * 
      length(dat07$var2) * 6,  
    geom = "area", alpha = 0.5, fill = "linen", col = "black") +
  labs(title = "Histogram of var2", y = "")
plot2C <- ggplot(dat07, aes(x = var2, y = "")) +
  geom_violin(fill = "plum") + geom_boxplot(width = 0.3) +
  stat_summary(fun = "mean", geom = "point", col = "red") +
  labs(title = "Boxplot of var2", y = "", x = "var2")

plot2A + (plot2B / plot2C + plot_layout(heights = c(2,1))) 
```

## Three-Plot Summary for `var03`

```{r}
#| echo: false

plot3A <- ggplot(dat07, aes(sample = var3)) +
  geom_qq(shape = 21, fill = "coral") + geom_qq_line(col = "red") + 
  labs(title = "Normal Q-Q plot of var3", y = "var3", x = "Standard Normal")
plot3B <- ggplot(dat07, aes(x = var3)) +
  geom_histogram(binwidth = 6, col = "black", fill = "coral") + 
  stat_function(fun = function(x)
    dnorm(x, mean = mean(dat07$var3, na.rm = TRUE),
          sd = sd(dat07$var3,na.rm = TRUE)) * 
      length(dat07$var3) * 6,  
    geom = "area", alpha = 0.5, fill = "linen", col = "black") +
  labs(title = "Histogram of var3", y = "")
plot3C <- ggplot(dat07, aes(x = var3, y = "")) +
  geom_violin(fill = "coral") + geom_boxplot(width = 0.3) +
  stat_summary(fun = "mean", geom = "point", col = "red") +
  labs(title = "Boxplot of var3", y = "", x = "var3")

plot3A + (plot3B / plot3C + plot_layout(heights = c(2,1))) 
```

## Three-Plot Summary for `var04`

```{r}
#| echo: false

plot4A <- ggplot(dat07, aes(sample = var4)) +
  geom_qq(shape = 21, fill = "aquamarine") + geom_qq_line(col = "red") + 
  labs(title = "Normal Q-Q plot of var4", y = "var4", x = "Standard Normal")
plot4B <- ggplot(dat07, aes(x = var4)) +
  geom_histogram(binwidth = 4, col = "black", fill = "aquamarine") + 
  stat_function(fun = function(x)
    dnorm(x, mean = mean(dat07$var4, na.rm = TRUE),
          sd = sd(dat07$var4,na.rm = TRUE)) * 
      length(dat07$var4) * 4,  
    geom = "area", alpha = 0.5, fill = "linen", col = "black") +
  labs(title = "Histogram of var4", y = "")
plot4C <- ggplot(dat07, aes(x = var4, y = "")) +
  geom_violin(fill = "aquamarine") + geom_boxplot(width = 0.3) +
  stat_summary(fun = "mean", geom = "point", col = "red") +
  labs(title = "Boxplot of var4", y = "", x = "var3")

plot4A + (plot4B / plot4C + plot_layout(heights = c(2,1))) 
```

## Three-Plot Summary for `var05`

```{r}
#| echo: false

plot5A <- ggplot(dat07, aes(sample = var5)) +
  geom_qq(shape = 21, fill = "slateblue") + geom_qq_line(col = "red") + 
  labs(title = "Normal Q-Q plot of var5", y = "var5", x = "Standard Normal")
plot5B <- ggplot(dat07, aes(x = var5)) +
  geom_histogram(binwidth = 6, col = "black", fill = "slateblue") + 
  stat_function(fun = function(x)
    dnorm(x, mean = mean(dat07$var5, na.rm = TRUE),
          sd = sd(dat07$var5,na.rm = TRUE)) * 
      length(dat07$var5) * 6,  
    geom = "area", alpha = 0.5, fill = "linen", col = "black") +
  labs(title = "Histogram of var5", y = "")
plot5C <- ggplot(dat07, aes(x = var5, y = "")) +
  geom_violin(fill = "slateblue", alpha = 0.5) + geom_boxplot(width = 0.3) +
  stat_summary(fun = "mean", geom = "point", col = "red") +
  labs(title = "Boxplot of var5", y = "", x = "var5")

plot5A + (plot5B / plot5C + plot_layout(heights = c(2,1))) 
```

## Three-Plot Summary for `var06`

```{r}
#| echo: false

plot6A <- ggplot(dat07, aes(sample = var6)) +
  geom_qq(shape = 21, fill = "burlywood") + geom_qq_line(col = "red") + 
  labs(title = "Normal Q-Q plot of var6", y = "var6", x = "Standard Normal")
plot6B <- ggplot(dat07, aes(x = var6)) +
  geom_histogram(binwidth = 6, col = "black", fill = "burlywood") + 
  stat_function(fun = function(x)
    dnorm(x, mean = mean(dat07$var6, na.rm = TRUE),
          sd = sd(dat07$var6,na.rm = TRUE)) * 
      length(dat07$var6) * 6,  
    geom = "area", alpha = 0.5, fill = "linen", col = "black") +
  labs(title = "Histogram of var6", y = "")
plot6C <- ggplot(dat07, aes(x = var6, y = "")) +
  geom_violin(fill = "burlywood") + geom_boxplot(width = 0.3) +
  stat_summary(fun = "mean", geom = "point", col = "red") +
  labs(title = "Boxplot of var6", y = "", x = "var6")

plot6A + (plot6B / plot6C + plot_layout(heights = c(2,1))) 
```

## New Data: Ohio's 88 Counties

Adult obesity rates (% of adult county residents reporting body-mass index of 30 $kg/m^2$ or higher) from County Health Rankings Reports in 2014 and 2024.

- CHR 2024 reports **2021** data from the Behavioral Risk Factor Surveillance System.
- CHR 2014 reports **2010** data from the National Center for Chronic Disease Prevention and Health Promotion, Division of Diabetes Translation.

## Obesity in Ohio Counties

```{r}
#| echo: true

ohio88 <- read_csv("c07/data/ohio88.csv", show_col_types = FALSE) |>
  janitor::clean_names() |> 
  mutate(obesity_diff = obesity_2024 - obesity_2014) |>
  select(county, obesity_2024, obesity_2014, obesity_diff)

ohio88 |> filter(county %in% 
                  c("Cuyahoga", "Geauga", "Lake", "Lorain", 
                    "Medina", "Portage", "Summit"))
```

## Our Question

What can we say about the mean of the difference between the county-level obesity rates reported in 2024 and those reported in 2014?

- What is the outcome of interest?
- What two groups are we comparing?
- Do we have paired or independent samples here? 

## Numerical Summaries

```{r}
#| echo: true

res1 <- ohio88 |> reframe(lovedist(obesity_2024)) |> mutate(var = "2024")
res2 <- ohio88 |> reframe(lovedist(obesity_2014)) |> mutate(var = "2014")
res3 <- ohio88 |> reframe(lovedist(obesity_diff)) |> mutate(var = "diffs")

rbind(res1, res2, res3) |> relocate(var) |>
  kable(digits = c(0, 0, 0, 2, 2, 2, 2, 1, 1, 1, 1)) |>
  kable_styling(font_size = 28)
```


## Plotting Paired Differences

```{r}
#| echo: true
#| output-location: slide

p1 <- ggplot(ohio88, aes(sample = obesity_diff)) +
  geom_qq() +
  geom_qq_line(col = "red") +
  labs(y = "2024 - 2014 difference", 
       x = "Standard Normal Distribution",
    title = "Changes in Adult Obesity, 2024-2014")

bw <- 1 # specify width of bins in histogram

p2 <- ggplot(ohio88, aes(obesity_diff)) +
  geom_histogram(binwidth = bw, fill = "black", col = "yellow" ) +
  stat_function(fun = function(x) {
    dnorm(x, mean = mean(ohio88$obesity_diff, na.rm = TRUE),
          sd = sd(ohio88$obesity_diff, na.rm = TRUE) ) *
        length(ohio88$obesity_diff) * bw},
    geom = "area", alpha = 0.5, fill = "lightblue", col = "blue") +
  labs(x = "2024 - 2014 difference", y = "", 
       title = "Histogram of 88 Ohio Counties")

p3 <- ggplot(ohio88, aes(x = obesity_diff, y = "")) +
  geom_violin(fill = "cornsilk") +
  geom_boxplot(width = 0.2) +
  stat_summary(fun = mean, geom = "point", shape = 16, col = "red") +
  labs(y = "", x = "2024 - 2014 difference", 
       title = "Boxplot with Violin: 88 Ohio Counties")

p1 + (p2 / p3 + plot_layout(heights = c(2, 1))) +
  plot_annotation(title = "2024 - 2014 Difference in CHR Adult Obesity Rates")
```

## Our paired differences

```{r}
#| echo: true

ohio88 |> tabyl(obesity_diff) |> adorn_pct_formatting()
```


## Our 88 Paired Differences

```{r}
#| echo: true
ohio88 |> reframe(lovedist(obesity_diff))
```

What can we say about the distribution of paired differences in our sample?

- Do we see meaningful skew?
- Do we see meaningfully heavy or light tails?
- Which comparison approach should we use?

## OLS/t-based approach

```{r}
#| echo: true
m2a <- lm(obesity_diff ~ 1, data = ohio88)

model_parameters(m2a, conf.level = 0.95)
```

```{r}
#| echo: true
m2b <- t.test(ohio88$obesity_diff, conf.level = 0.95)
model_parameters(m2b)
```

## Bayesian Linear Model

```{r}
#| echo: true

set.seed(43101)
m3 <- stan_glm(obesity_diff ~ 1, data = ohio88, refresh = 0)
model_parameters(m3, conf.level = 0.95)
```


## Bootstrap approach

```{r}
#| echo: true

set.seed(432071)
boot.t.test(ohio88$obesity_diff, R = 2000, conf.level = 0.95)
```

## Our estimates

Approach | Estimate | 95% confidence interval
:--------: | :--------: | :-----------------------:
OLS / t | 9.40 | (8.77, 10.02)
Bayes | 9.39 | (8.77, 10.03)
Bootstrap | 9.40 | (8.78, 9.99)

- Which of these should we choose?
  - Or does it not matter much?
- Once we've made a choice, how can we describe the result properly?

## Choosing OLS for Paired t

*Paired t result: Estimate = 9.40, 95% CI (8.77, 10.02)*

- Our best point estimate for the true difference (2024-2014 reports on adult obesity) is 9.40 percentage points, based on our sample of 88 Ohio counties.
- The confidence interval reflects imprecision in the population estimate, based on assuming that the counties in our sample are selected at random from the population of interest, and that the assumptions of our paired t procedure hold.

## Interpreting the Confidence Interval

*Paired t result: Estimate = 9.40, 95% CI (8.77, 10.02)*

When we generalize beyond these counties to the population they were **selected at random** from, then our data are **compatible** (at the 95% confidence level) with population mean differences in adult obesity rates between 8.77 and 10.02 percentage points, depending on the **assumptions of our linear model** (paired t test) being correct.

## Paired Samples Strategy (1)

Suppose we want to estimate an uncertainty interval for the mean of a set of paired differences.

- Calculate the paired differences, then plot them.
- If the sample data are well described as "Normal", then use the OLS / paired t procedure to obtain a confidence interval, or a Bayesian model to obtain a credible interval.

## Paired Samples Strategy (2)

- If the sample data are best described as "symmetric but with outliers", then use the bootstrap to obtain a confidence interval, although an OLS or Bayesian result may be quite similar.
- If the sample data are best described as "substantially skewed", then consider whether you actually want to summarize with the mean, and consider whether a transformation might be helpful.

# Standing Break

## New Data: `dm464` from Classes 5-6

```{r}
#| echo: true
dm464 <- read_xlsx("c07/data/dm464.xlsx") |>
  janitor::clean_names() |>
  select(id_code, ldl = ldl_start, statin = statin_start) |>
  mutate(across(where(is.character), as_factor)) |>
  mutate(id_code = as.character(id_code)) |>
  mutate(stat_f = fct_recode(factor(statin), "Statin" = "1", "No" = "0"))

head(dm464)
```

## What is our goal?

Goal: 90% uncertainty interval for difference in average LDL between Statin and No Statin groups.

```{r}
#| echo: true

dm464 |> group_by(stat_f) |> reframe(lovedist(ldl)) |> 
  kable(digits = c(0, 0, 0, 2, 2, 2, 2, 0, 0, 0, 0)) |>
  kable_styling(font_size = 24)
```

::: {.incremental}
- What are we comparing? What is our outcome? How many group means?
- Are these paired or independent samples?
:::

## Estimate 1: Pooled t procedure

Pooled t procedure using `lm` (ordinary least squares)

```{r}
#| echo: true
fit1 <- lm(ldl ~ stat_f, data = dm464)
model_parameters(fit1, ci = 0.90) |> 
  kable(digits = 2) |> kable_styling(font_size = 24)
```

### Assumptions for Pooled t procedure

1. Random sample from statin and no statin populations
2. Each sample comes from a Normal population
3. Equal population variances in the two statin groups

## Interpreting the Pooled t result (1/2)

*Pooled t result: Estimate = 7.17, 90% CI (0.02, 14.32)*

- Our best point estimate for the true difference (Statin - No Statin) is 7.17 mg/dl in LDL cholesterol at baseline, based on our sample of 367 Statin and 97 No statin subjects with diabetes.
- The confidence interval reflects imprecision in the population estimate, based on assuming that the people in our sample are selected at random from the population of interest, *and* that the assumptions of our pooled t procedure hold.

## Interpreting the Pooled t result (2/2)

When we generalize beyond these 464 subjects with diabetes to the population they were **selected at random** from, then our data are **compatible** (at the 90% confidence level) with population mean differences in LDL between 0.02 and 14.32 mg/dl, depending on the **assumptions of our linear model** (paired t test) being correct.

## Estimate 2: Bayesian linear model

If we use the default (weakly informative) prior, then the Bayesian model has essentially the same assumptions as the pooled t procedure:

```{r}
#| echo: true
set.seed(4312)
fit2 <- stan_glm(ldl ~ stat_f, data = dm464, refresh = 0)
model_parameters(fit2, ci = 0.90) |> 
  kable(digits = 2) |> kable_styling(font_size = 24)
```

## Estimate 3: Welch's t procedure

Welch's t procedure drops requirement of equal population variances

```{r}
#| echo: true
fit3 <- t.test(ldl ~ stat_f, data = dm464)
model_parameters(fit3, ci = 0.90)
```

## Estimate 4: Bootstrap with pooling

Bootstrap (percentile) procedure (with pooling) maintains assumption of equal population variances but does not require Normally distributed populations

```{r}
#| echo: true
set.seed(43104)
boot.t.test(ldl ~ stat_f, var.equal = TRUE, R = 2000,
  data = dm464, conf.level = 0.90)
```

## Estimate 5: Bootstrap: unpooled

Bootstrap (percentile) procedure (without pooling) drops assumptions of equal population variances and of Normally distributed populations

```{r}
#| echo: true
set.seed(43105)
boot.t.test(ldl ~ stat_f, var.equal = FALSE, R = 2000,
  data = dm464, conf.level = 0.90)
```

## Results So Far{.smaller}

- All methods assume random samples
- Est. = point estimate of Statin - No Statin group mean difference

Method | Est. | 90% Interval | Assumptions
------ | --------- | -------------- | ----------------------------
Pooled t | 7.17 | (0.02, 14.32) | Normality, equal variances
Bayesian | 7.16 | (0.17, 14.65) | Same as pooled t, basically
Welch t | 7.17 | (-0.36, 14.70) | Normality only
Boot (pooled) | 7.25 | (0.18, 14.28) | Equal variances only
Boot (Welch) | 7.06 | (0.87, 12.99) | Neither

- So what do we need to look for in our data?

## LDL by Statin prescription status

Comparison Boxplot with Violins

```{r}
#| echo: true
#| output-location: slide

ggplot(dm464, aes(x = ldl, y = stat_f)) +
  geom_violin() +
  geom_boxplot(aes(fill = stat_f), width = 0.3) +
  stat_summary(fun = mean, geom = "point", 
               shape = 18, size = 4, col = "black") +
  guides(fill = "none") +
  labs(y = "Statin prescription", x = "LDL Cholesterol (mg/dl)",
       title = "LDL by Statin prescription")
```

## LDL by Statin prescription status

How about a pair of Normal Q-Q plots?

```{r}
#| echo: true
#| output-location: slide

p1 <- dm464 |>
  filter(stat_f == "Statin") |>
  ggplot(aes(sample = ldl)) +
  geom_qq() + geom_qq_line(col = "red") +
  labs(title = "Statin group", y = "LDL Cholesterol (mg/dl)", 
        x = "Standard Normal")

p2 <- dm464 |>
  filter(stat_f == "No") |>
  ggplot(aes(sample = ldl)) +
  geom_qq() + geom_qq_line(col = "red") +
  labs(title = "No Statin group", y = "LDL Cholesterol (mg/dl)", 
        x = "Standard Normal")

p1 + p2 +
  plot_annotation(title = "Normal Q-Q plots of LDL by Statin group")
```

## LDL by Statin prescription status

A look at some numerical summaries...

```{r}
#| echo: true

dm464 |> group_by(stat_f) |> reframe(lovedist(ldl)) |> 
  kable(digits = 2) |> kable_styling(font_size = 28)
```

- skew~1~ for "Statin" = (103.43 - 93)/39.51 = 0.26
- skew~1~ for "No" = (96.26 - 91)/31.63 = 0.17

Our rough cutoff for substantial right skew ~ 0.20.

## What can we do about the skew?

- Could we build a confidence interval for medians instead of means?
  - Yes, with the bootstrap, for example see [Section 6.9.2](https://thomaselove.github.io/431-book/06_twogroups.html#bootstrap-ci-for-medians) of my text.

- Could we transform the data in a non-linear way to better match assumptions?
  - Yes, with the help of **Tukey's ladder** of power transformations.

## Tukey's Ladder

Power ($\lambda$) | -2 | -1 | -0.5 | 0 | 0.5 | *1* | 2 | 3
------------: | ---: | ---: | ---: | -----: | ---: | ---: | ---: | ---:
Transform | $\frac{1}{y^2}$ | $\frac{1}{y}$ | $\frac{1}{\sqrt{y}}$ | $\log{y}$ | $\sqrt{y}$ | *y* | $y^2$ | $y^3$

- Start at no transformation ($\lambda$ = 1)
  - Try moving in one direction, searching for a better result (more "Normal" distribution)
  - Keep going until you start turning in the wrong direction.

## Tukey's Ladder

Power ($\lambda$) | -2 | -1 | -0.5 | 0 | 0.5 | 1 | 2 | 3
------------: | ---: | ---: | ---: | -----: | ---: | ---: | ---: | ---:
Transform | $\frac{1}{y^2}$ | $\frac{1}{y}$ | $\frac{1}{\sqrt{y}}$ | $\log{y}$ | $\sqrt{y}$ | $y$ | $y^2$ | $y^3$

- These power transformations help us address skew, mainly. 
- They're not much help if the data are symmetric, but have heavy/light tails.
- Some transformations require all values to be positive. If they aren't, we can add a small amount to each observation until they are.

## LDL (untransformed) by statin

```{r}
ggplot(dm464, aes(x = ldl, y = stat_f)) +
  geom_violin() +
  geom_boxplot(aes(fill = stat_f), width = 0.3) +
  stat_summary(fun = mean, geom = "point", 
               shape = 18, size = 4, col = "black") +
  guides(fill = "none") +
  labs(y = "Statin prescription", x = "LDL Cholesterol (mg/dl)",
       title = "LDL by Statin prescription",
       subtitle = "Each group is a bit skewed to the right")
```

- Let's try $\lambda = 2$, **squaring** the LDL values.

## LDL squared, by statin

```{r}
#| echo: true
#| output-location: slide

ggplot(dm464, aes(x = ldl^2, y = stat_f)) +
  geom_violin() +
  geom_boxplot(aes(fill = stat_f), width = 0.3) +
  stat_summary(fun = mean, geom = "point", 
               shape = 18, size = 4, col = "black") +
  guides(fill = "none") +
  labs(y = "Statin prescription", x = "LDL^2",
       title = "LDL^2 by Statin prescription")
```

- Is this an improvement?



## Square root of LDL?

```{r}
#| echo: true
#| output-location: slide

ggplot(dm464, aes(x = sqrt(ldl), y = stat_f)) +
  geom_violin() +
  geom_boxplot(aes(fill = stat_f), width = 0.3) +
  stat_summary(fun = mean, geom = "point", 
               shape = 18, size = 4, col = "black") +
  guides(fill = "none") +
  labs(y = "Statin prescription", x = "sqrt(LDL)",
       title = "Square Root of LDL by Statin prescription")
```

## Logarithm of LDL?

```{r}
#| echo: true
#| output-location: slide

ggplot(dm464, aes(x = log(ldl), y = stat_f)) +
  geom_violin() +
  geom_boxplot(aes(fill = stat_f), width = 0.3) +
  stat_summary(fun = mean, geom = "point", 
               shape = 18, size = 4, col = "black") +
  guides(fill = "none") +
  labs(y = "Statin prescription", x = "log(LDL)",
       title = "Logarithm of LDL by Statin prescription")
```

## Inverse of LDL?

```{r}
#| echo: true
#| output-location: slide

ggplot(dm464, aes(x = (1/ldl), y = stat_f)) +
  geom_violin() +
  geom_boxplot(aes(fill = stat_f), width = 0.3) +
  stat_summary(fun = mean, geom = "point", 
               shape = 18, size = 4, col = "black") +
  guides(fill = "none") +
  labs(y = "Statin prescription", x = "1/LDL",
       title = "Inverse of LDL by Statin prescription")
```

## Which Transformation to Choose?

Box-Cox approach: can we get a suggested "power" to use when transforming our outcome?

- Specify the model then apply `boxCox` from `car` package.
- Here, all values of our outcome (LDL) are strictly positive. 
  - If not, we'd have to add a constant so that they were.

### Tukey's Ladder

Power ($\lambda$) | -2 | -1 | -0.5 | 0 | 0.5 | 1 | 2 | 3
------------: | ---: | ---: | ---: | -----: | ---: | ---: | ---: | ---:
Transform | $\frac{1}{y^2}$ | $\frac{1}{y}$ | $\frac{1}{\sqrt{y}}$ | $\log{y}$ | $\sqrt{y}$ | $y$ | $y^2$ | $y^3$

## Box-Cox Suggestion?

```{r}
#| echo: true

fit6 <- lm(ldl ~ stat_f, data = dm464)
boxCox(fit6)
```

## Logarithm of LDL?

```{r}
#| echo: true
#| output-location: slide

p1 <- dm464 |>
  filter(stat_f == "Statin") |>
  ggplot(aes(sample = log(ldl))) +
  geom_qq() + geom_qq_line(col = "red") +
  labs(title = "Statin group", y = "log(LDL Cholesterol)", 
        x = "Standard Normal")

p2 <- dm464 |>
  filter(stat_f == "No") |>
  ggplot(aes(sample = log(ldl))) +
  geom_qq() + geom_qq_line(col = "red") +
  labs(title = "No Statin group", y = "log(LDL Cholesterol)", 
        x = "Standard Normal")

p1 + p2 +
  plot_annotation(title = "Normal Q-Q plots of log(LDL) by Statin group")
```


## Using the log transformation

```{r}
#| echo: true
#| message: true
fit7 <- lm(log(ldl) ~ stat_f, data = dm464)
model_parameters(fit7, ci = 0.90)  
```

- How do we interpret these results?
    - Should we exponentiate?

## Using the log transformation

```{r}
#| echo: true
fit7 <- lm(log(ldl) ~ stat_f, data = dm464)
model_parameters(fit7, ci = 0.90)  
```

- Subjects with statin prescriptions have, on average, 0.05 higher *log(LDL)* than subjects without a statin prescription. 
- The 90% confidence interval for the Statin - No Statin difference in *log (LDL)* is (-0.02, 0.12).

## Exponentiating the Coefficients

```{r}
#| echo: true
fit7 <- lm(log(ldl) ~ stat_f, data = dm464)
model_parameters(fit7, ci = 0.90, exponentiate = TRUE)  
```

- After exponentiating, the point estimate for the comparison is 1.05, with 90% CI (0.98, 1.13).
- Subjects with statin prescriptions have, on average, 5% higher LDL than do subjects without such a prescription.
  - How should we interpret the confidence interval?

## Interpreting the Result Properly

- In our sample, subjects *with* a statin prescription have 5% higher LDL, on average, than do subjects *without*. 
- When we generalize beyond these 464 subjects with diabetes to the population they were **selected at random** from, then our data are **compatible** (at the 90% confidence level) with population LDL values for statin users that are between 98% and 113% as large as non-statin users, depending on the **assumptions of our linear model** being correct after our log transformation of the outcome.

## Independent Samples Strategy (1)

Suppose we want to estimate an uncertainty interval for the difference in means across two groups. To begin, plot the data from each sample.

- If the sample data in each group are well described as "Normal", then use the OLS / pooled t procedure to obtain a confidence interval, or a Bayesian model to obtain a credible interval.
    - If the sample sizes are not the same, and the sample variances aren't close to each other, consider using a Welch t procedure.

## Independent Samples Strategy (2)

- If either sample's data are "symmetric but with outliers", then consider using the bootstrap to obtain a confidence interval, although an OLS or Bayesian result may also be reasonable.
- If either sample's data are best described as "substantially skewed", then consider whether you actually want to summarize with the difference in means, and consider whether a transformation might be helpful.

## Session Information

```{r}
#| echo: true

xfun::session_info()
```

