---
title: "431 Class 09"
author: Thomas E. Love, Ph.D.
date: "2026-09-22"
format:
  revealjs: 
    theme: dracula
    embed-resources: true
    self-contained: true
    slide-number: true
    footnotes-hover: true
    preview-links: auto
    date-format: iso
    footer: "431 Class 09 | 2026-09-22 | <https://thomaselove.github.io/431-2026/>"
---

## Today's Agenda

- New Data on European Family Care Givers
- Comparing Working Hours by Education Level
  - One-Way Analysis of Variance
  - Considering an Outcome Transformation in ANOVA
- Correlation and Regression to study burden
  - Fitting a linear model with `lm()`
  - Checking its performance and fit
- **1:45 PM**: Favorite Movies: Session 2

## Load packages and set theme

```{r}
#| echo: true
#| message: false

knitr::opts_chunk$set(comment = NA)

source("c09/data/Love-431.R") # for the lovedist() function

library(janitor)
library(patchwork)
library(DescTools)  ## post-hoc Bonferroni or Tukey HSD
library(naniar)     ## dealing with missing data
library(car)        ## for Box-Cox    
library(easystats)
library(tidyverse)

theme_set(theme_lucid()) 
```

## Get the data

[EUROFAMCARE project](https://www.uke.de/extern/eurofamcare/): a survey on the situation of family carers of older people in Europe.

```{r}
#| echo: true
ecare_raw <- read_csv("c09/data/eurocare.csv", show_col_types = FALSE) |>
  janitor::clean_names() |>
  mutate(across(where(is.character), as_factor)) |>
  mutate(subject = as.character(subject)) |>
  select(subject, hours, c_educ = c_education, neg_impact, pos_impact)

dim(ecare_raw)
head(ecare_raw, 3)
```

## Variable Descriptions

Variable | Description
---------- | :------------------------------------------
`subject` | carer (care giver) associated with this family
`hours` | weekly hours carer spends caring for elder
`c_educ` | carer education level (High, Medium, Low)
`neg_impact` | COPE Index: carer's burden: negative impact
`pos_impact` | COPE Index: carer's burden: positive impact

More on these burden scores later.

## Missingness Assessment

```{r}
#| echo: true

miss_var_summary(ecare_raw)
miss_case_table(ecare_raw)
```

## Eliminating Missing Values

Today, we'll filter to complete cases across the whole data set.

```{r}
#| echo: true

ecare <- ecare_raw |> drop_na()
summary(ecare)
```

Complete case analyses can create problems, but for today, we'll ignore them.

## `c_educ` is a factor

```{r}
#| echo: true
ecare |> tabyl(c_educ)
```

Put this factor into a more sensible order with `fct_relevel()`

```{r}
#| echo: true
ecare <- ecare |> 
  mutate(c_educ = 
           fct_relevel(c_educ, "High", "Middle", "Low"))
```

## Automated Codebook

We don't need to see the subjects list.

```{r}
#| echo: true
data_codebook(ecare |> select(!subject))
```

## How do mean `hours` vary by `c_educ`?

```{r}
#| echo: true
#| output-location: slide

ggplot(ecare, aes(y = hours, x = c_educ)) +
  geom_violin(aes(fill = c_educ)) + 
  geom_boxplot(width = 0.2) +
  stat_summary(fun = mean, geom = "point", 
               shape = 16, size = 3, col = "red") +
  scale_fill_viridis_d(alpha = 0.3) +
  guides(fill = "none")
```

## How do `hours` vary by `c_educ`? (1/2)

```{r}
#| echo: true

ecare |> group_by(c_educ) |> reframe(lovedist(hours))
```

- Sample mean `hours` are somewhat smaller for carers with higher levels of education. 
  - Does these mean differences seem meaningful?
  - Does each group's data follow a Normal distribution?

## Do `hours` vary by `c_educ`? (2/2)

```{r}
#| echo: true

ecare |> group_by(c_educ) |> reframe(lovedist(hours))
```

- What else might we consider here?
  - Do we have a balanced design?
  - Are sample variances very different from each other?

## ANOVA model for this comparison

```{r}
#| echo: true

fit1 <- lm(hours ~ c_educ, data = ecare)
anova(fit1)
model_parameters(fit1, ci = 0.95)
```

## Plot means with Bonferroni 95% CIs

- Tukey is less appealing in light of unbalanced design.

```{r}
#| echo: true

par(mar=c(3,10,3,3), las = 1)
plot(PostHocTest(aov(fit1), method = "bonferroni", conf.level = 0.95))
```

## Does model `fit1` work well?

```{r}
#| echo: true

plot(fit1, which = 2)
```

## Could we transform the outcome?

```{r}
#| echo: true

boxCox(fit1)
```

## Impact of log transformation

```{r}
#| echo: true

ecare |> group_by(c_educ) |> reframe(lovedist(log(hours)))
```

```{r}
#| echo: true
#| output-location: slide

ggplot(ecare, aes(y = log(hours), x = c_educ)) +
  geom_violin(aes(fill = c_educ)) + 
  geom_boxplot(width = 0.2) +
  stat_summary(fun = mean, geom = "point", 
               shape = 16, size = 3, col = "red") +
  scale_fill_viridis_d(alpha = 0.3) +
  guides(fill = "none")
```

## ANOVA after Transforming `hours`

```{r}
#| echo: true

fit2 <- lm(log(hours) ~ c_educ, data = ecare)

anova(fit2)
```

- What's the conclusion here?

## Bonferroni 95% CIs for log(hours)

```{r}
#| echo: true

PostHocTest(aov(fit2), method = "bonferroni", conf.level = 0.95)
```

- Variable predicted is actually **log(hours)**

## Bonferroni 95% CIs for log(hours)

```{r}
#| echo: true

par(mar=c(3,10,3,3), las = 1)
plot(PostHocTest(aov(fit2), method = "bonferroni", conf.level = 0.95))
```

## Normality still a problem?

```{r}
#| echo: true

plot(fit2, which = 2)
```

## Kruskal-Wallis ANOVA

If the data were severely skewed, you might try:

```{r}
#| echo: true

kruskal.test(hours ~ c_educ, data = ecare)
```

- $H_0$: The four education groups have the same center to their `hours` distributions.
- $H_A$: At least one group has a shifted distribution, with a different center to its `hours`.

## K Samples: Comparing Means (1)

1. What is the outcome under study?
2. What are the ($K \geq 2$) treatment/exposure groups?
3. Were the data in fact collected using independent samples?
4. Are the data random samples from the population(s) of interest? Or is there at least a reasonable argument for generalizing from the samples to the population(s)?
5. What confidence level do we require?

## K Samples: Comparing Means (2)

6. Are we doing one-sided or two-sided testing? (usually 2-sided)
7. What does the distribution of each individual sample tell us about which inferential procedure to use?
8. Are there meaningful differences between population means?
9. Can we identify pairwise comparisons of means that show meaningful differences using an appropriate procedure that protects against Type I error expansion due to multiple comparisons?

# On to Linear Regression

## Association of Two Quantities

How well can we predict a carer’s negative impact score on the basis of their positive impact score?

- The two scales are described further on the next slide.

```{r}
#| echo: true

ecare |> reframe(lovedist(neg_impact)) 
ecare |> reframe(lovedist(pos_impact)) 
```


## Caregiver Burden Subscales

- `neg_impact` is higher for carers characterized by avoiding problems, focusing on negative emotions, or engaging in unhealthy behaviors, which are linked to increased psychological distress and other adverse effects.
  - scores in our data range from 7 to 28

- `pos_impact` is higher for carers characterized by adaptive coping strategies focused on finding positive meaning and growth or positive outcomes in stressful situations. 
  - scores in our data range from 5 to 16

## Histograms of burden scales

```{r}
plot1 <- ggplot(ecare, aes(x = neg_impact)) +
       geom_histogram(binwidth = 1, fill = "blue", col = "white") + 
       labs(title = "Negative Impact")

plot2 <- ggplot(ecare, aes(x = pos_impact)) +
       geom_histogram(binwidth = 1, fill = "forestgreen", col = "white") + 
       labs(title = "Positive Impact")

plot1 + plot2
```

## Scatterplot of 818 Carers

- We’ll plot each point with alpha = 0.3 so that if we have multiple overlapping points, the plot will look darker there.

```{r}
#| echo: true
#| output-location: slide

ggplot(ecare, aes(x = pos_impact, y = neg_impact)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", 
              formula = y ~ x, se = FALSE, col = "blue") +
  geom_smooth(method = "lm", 
              formula = y ~ x, se = TRUE, col = "red") +
  labs(title = "EUROFAMCARE data for 818 subjects",
       x = "Positive Impact Score", 
       y = "Negative Impact Score",
       caption = "Linear model in red, loess smooth in blue")
```

## Jittered Scatterplot

```{r}
#| echo: false

ggplot(ecare, aes(x = pos_impact, y = neg_impact)) +
  geom_jitter() +
  geom_smooth(method = "loess", 
              formula = y ~ x, se = FALSE, col = "blue") +
  geom_smooth(method = "lm", 
              formula = y ~ x, se = TRUE, col = "red") +
  labs(title = "EUROFAMCARE data for 818 subjects",
       x = "Positive Impact Score", 
       y = "Negative Impact Score",
       caption = "Linear model in red, loess smooth in blue")
```

## Pearson Correlation Coefficent

- Pearson r ranges from -1 (perfect inverse relationship: as x goes up, y goes down) to +1 (perfect direct relationship: as x goes up, y goes up.)
- Pearson r measures only a *linear* association.

```{r}
#| echo: true

ecare |> select(neg_impact, pos_impact) |> correlation()
```

## Considering the Correlation

```{r}
#| echo: true

ecare |> select(neg_impact, pos_impact) |> correlation()
```

- Is the direction of this correlation surprising?
- How about its magnitude?

## Fit a linear model with OLS

Ordinary Least Squares used to fit a linear model...

```{r}
#| echo: true

fit3 <- lm(neg_impact ~ pos_impact, data = ecare)
model_parameters(fit3, ci = 0.97, pretty_names = FALSE)
```

Note that we're creating **97%** confidence intervals here.

## Interpreting the intercept

```{r}
#| echo: true

model_parameters(fit3, ci = 0.97, pretty_names = FALSE)
```

- Our `fit3` model estimates the `neg_impact` value to be 20.69 if `pos_impact` = 0.
- All of our `pos_impact` values are between 5 and 16, so we don't have any useful information about what happens when `pos_impact` = 0, and thus the intercept isn't actually useful.

## Interpret slope estimate

```{r}
#| echo: true

model_parameters(fit3, ci = 0.97, pretty_names = FALSE)
```

- Suppose Harry's `pos_impact` is one point larger than Sally's. Our `fit3` model predicts the `neg_impact` for Harry to be 0.71 points less than that of Sally, on average.
- Of course, this assumes Harry and Sally are similar to the subjects in `ecare` and that our linear model `fit3` is correct.

## Interpret slope's 97% CI

```{r}
#| echo: true

model_parameters(fit3, ci = 0.97, pretty_names = FALSE)
```

- When we generalize beyond these 818 study participants to the population they were selected at random from, then our data are compatible (at the 97% confidence level) with population slopes between -0.83 and -0.59, depending on the assumptions of our model `fit3` being correct.

## Performance of this model?

```{r}
#| echo: true

model_performance(fit3)
```

- $R^2$ tells us model `fit3` accounts for 16.9% of the variation in `neg_impact` we see in these 818 subjects.
- $R^2$ = 0.169 is the square of the Pearson correlation (-0.41) of `neg_impact` and `pos_impact`. 
- $\eta^2$ in ANOVA = this $R^2$.

::: aside

Adjusted $R^2$ will be useful when we have multiple predictors.

:::

## Performance of this model?

```{r}
#| echo: true

model_performance(fit3)
```

The model `fit3` assumes that our errors (residuals) come from a Normal distribution with mean 0 and standard deviation `Sigma` ($\sigma) = 3.546.

- About 68% of our predicted `neg_impact` values should be within $\pm 3.546$ points of the correct value, and 95% should be within $\pm 2(3.546)$, or $\pm 7.092$ points. 

# Favorite Movies Data

## Session Information

```{r}
#| echo: true

xfun::session_info()
```


