---
title: "431 Class 06"
author: Thomas E. Love, Ph.D.
date: "2026-09-10"
format:
  revealjs: 
    theme: dracula
    embed-resources: true
    self-contained: true
    slide-number: true
    footnotes-hover: true
    preview-links: auto
    date-format: iso
    footer: "431 Class 06 | 2026-09-10 | <https://thomaselove.github.io/431-2026/>"
---

## Today's Agenda

- Assumptions of our uncertainty intervals for means
  - for one sample / paired differences
  - for two independent samples
- Transforming an Outcome (see [Chapter 7](https://thomaselove.github.io/431-book/))
  - The importance of power transformations
  - Tukey's ladder and the Box-Cox plot
  - Back-transforming predictions, not coefficients
  
::: aside

The last few slides describe a worked example for self-study.

:::

## Load packages and set theme

```{r}
#| echo: true
#| message: false

knitr::opts_chunk$set(comment = NA)
source("c06/data/Love-431.R") # for the lovedist() function

library(janitor)
library(patchwork)
library(readxl)
library(knitr)
library(kableExtra)
library(MKinfer)
library(rstanarm)
library(car)         ## new today
library(infer)       ## new today
library(easystats)
library(tidyverse)

theme_set(theme_bw())
```

## Our data: from `dm464` again

```{r}
#| echo: true
class06 <- read_xlsx("c06/data/dm464.xlsx") |>
  janitor::clean_names() |>
  select(id_code, age = age_start, sex, home = home_start,
          a1c_base = a1c_start, a1c_end, ldl = ldl_start, 
          statin = statin_start) |>
  mutate(across(where(is.character), as_factor)) |>
  mutate(id_code = as.character(id_code)) |>
  mutate(stat_f = as_factor(statin),
         stat_f = fct_recode(stat_f, "Statin" = "1", "No" = "0"))
```

- Read in the Excel data, then clean names and select (and rename some) variables.
- Change all character variables to factors, except `id_code`.
- Create a "Statin"/"No" factor for the 1/0 data in `statin`.

## The `class06` data

```{r}
#| echo: true

class06
```

```{r}
write_rds(class06, "c06/data/class06.Rds")
```

# Uncertainty Interval Assumptions in One Sample / Paired Samples

## One Sample (maybe Paired Differences)

- All approaches assume our sample is a **random** sample (or at least a representative one) from the population of interest.
- Ordinary Least Squares model / Paired T test also assumes that the population of interest follows a Normal distribution, so that our sample data should look as though it were drawn from a Normal distribution.

## One Sample (maybe Paired Differences)

- Our Bayesian approach assumes a (weakly informative) prior distribution on the coefficient in our model.
- The bootstrap approach does not assume a Normal distribution for the population.

Main check in data: Normal distribution?

## Does a Normal model fit well?

**Don't** rely on hypothesis tests of whether data follow a Normal distribution. Instead, do we have...

1. A histogram that is symmetric and bell-shaped.
2. A boxplot where the box is symmetric around the median, as are the whiskers, without severe outliers.
3. A normal Q-Q plot that essentially falls on a straight line.
4. If in doubt, maybe compare mean and sd to median and MAD, and perhaps consider Empirical Rule.

## Difference in A1c (end - baseline)

```{r}
#| echo: true
#| output-location: slide

class06 <- class06 |> mutate(a1c_diff = a1c_end - a1c_base)

plot1 <- ggplot(class06, aes(sample = a1c_diff)) +
  geom_qq() + geom_qq_line(col = "red") +
  labs(y = "A1c difference (%)", x = "Standard Normal Distribution",
    title = "Normal Q-Q plot")

bw = 1 # specify width of bins in histogram

plot2 <- ggplot(class06, aes(x = a1c_diff)) +
  geom_histogram(binwidth = bw, fill = "black", col = "yellow") +
  stat_function(fun = function(x) 
    dnorm(x, mean = mean(class06$a1c_diff, na.rm = TRUE), 
          sd = sd(class06$a1c_diff, na.rm = TRUE)) * 
          length(class06$a1c_diff) * bw,
    geom = "area", alpha = 0.5, 
    fill = "lightblue", col = "blue") +
  labs(x = "A1c difference (%)", y = "Count",
       title = "Histogram & Normal Curve")

plot3 <- ggplot(class06, aes(x = a1c_diff, y = "")) +
  geom_violin(fill = "cornsilk") +
  geom_boxplot(width = 0.2) +
  stat_summary(fun = mean, geom = "point", shape = 16, col = "red") +
  labs(y = "", x = "A1c difference (%)", title = "Boxplot with Violin")

plot1 + 
  (plot2 / plot3 + plot_layout(heights = c(2,1))) + 
  plot_annotation("Are differences in Hemoglobin A1c Normally distributed?")
```

## OLS/t test vs. Bayes

```{r}
#| echo: true
fit1 <- lm(a1c_diff ~ 1, data = class06)
model_parameters(fit1, ci = 0.9) 

set.seed(43101)
fit2 <- stan_glm(a1c_diff ~ 1, data = class06, refresh = 0)
model_parameters(fit2, ci = 0.9)
```

- *OLS*: estimate: 0.33, 90% confidence interval (0.20, 0.47)
- *Bayes*: estimate: 0.33, 90% credible interval (0.20, 0.47)

## Bootstrapped t test

```{r}
#| echo: true
set.seed(43102)
boot.t.test(class06$a1c_diff, conf.level = 0.90, R = 2000)
```

## What to do about outliers? (1/3)

The paired differences here appear to be symmetric, but with outliers.

- This should push us towards a method which doesn't require the assumption of Normality.
  - Specifically, the bootstrap would likely be a better choice than a t test, although it didn't make a meaningful difference.
  - 90% uncertainty interval with OLS/t test was (0.20, 0.47) and the bootstrap 90% uncertainty interval was (0.20, 0.48).

## What to do about outliers? (2/3)

```{r}
#| echo: true
class06 |> reframe(lovedist(a1c_diff)) |>
  kable(digits = 2) |> kable_styling(font_size = 28)
```

- **skew~1~** = (mean - median)/sd = (0.33 - 0.2)/1.79 = 0.073

Could we transform these data in a non-linear way to better match assumptions?

- Not really, no. Most of the transformations we use regularly are designed to address skew in the data, rather than tail behavior.

## What to do about outliers? (3/3)
  
Should we build our interval around the median instead of the mean? 

- sample median = 0.2, sample mean = 0.33

```{r}
#| fig-height: 2.5
#| fig-width: 10
ggplot(class06, aes(x = a1c_diff, y = "")) +
  geom_violin(fill = "cornsilk") +
  geom_boxplot(width = 0.2) +
  stat_summary(fun = mean, geom = "point", shape = 16, col = "red") +
  labs(y = "", x = "A1c difference (%)", title = "Boxplot with Violin")
```

- Maybe, but does that actually address the issue here?

## Bootstrap 90% interval for Median

This is from the `infer` package...

```{r}
#| echo: true
set.seed(43103)
x_med <- class06 |> observe(response = a1c_diff, stat = "median")
res1 <- class06 |>
  specify(response = a1c_diff) |>
  generate(reps = 2000, type = "bootstrap") |>
  calculate(stat = "median") |>
  get_confidence_interval(level = 0.90, type = "percentile")

res1 <- res1 |> mutate(pt_est = x_med$stat) |>
  relocate(pt_est)
res1
```

## A Strategy for Paired Samples {.smaller}

Suppose we want to estimate an uncertainty interval for the mean of a set of paired differences.

- Calculate the paired differences, then plot them.
- If the sample data are well described as "Normal", then use the OLS / paired t procedure to obtain a confidence interval, or a Bayesian model to obtain a credible interval.
- If the sample data are best described as "symmetric but with outliers", then use the bootstrap to obtain a confidence interval, although an OLS or Bayesian result (with a weakly informative prior) may be quite similar.
- If the sample data are best described as "substantially skewed", then consider whether you actually want to summarize with the mean, and consider whether a transformation of the outcome might be helpful.

# Uncertainty Interval Assumptions in Independent Samples

## Two Independent Samples (1/2)

- All methods assume each of the two samples is a random sample (or at least representative) of its population.
- Ordinary Least Squares model / Pooled T test assumes:
    - **each** of the two populations of interest follows a Normal distribution, **and**
    - **either** the variance of those two populations is equal, so it makes sense to pool the standard deviations, **or** the sample sizes are equal (a balanced design)

## Two Independent Samples (2/2)    
    
- The Welch t test makes the Normality assumption, but not the assumption of  equal population variances.
- Our Bayesian approach assumes a (weakly informative) prior distribution on each of the coefficients (intercept and slope) of our model.
- The bootstrap approach does not assume a Normal distribution for the population, but can take advantage of an assumption of equal variances if it exists.

## LDL means by `statin` group (1/5)

**OLS / pooled t**

```{r}
#| echo: true
fit3 <- lm(ldl ~ stat_f, data = class06)
model_parameters(fit3, ci = 0.90)
```

## LDL means by `statin` group (2/5)

**Welch t (no pooling)**

```{r}
#| echo: true
fit4 <- t.test(ldl ~ stat_f, data = class06, 
               var.equal = FALSE, conf.level = 0.90)
model_parameters(fit4, ci = 0.90)  
```

## LDL means by `statin` group (3/5)

**Bayesian linear model**

```{r}
#| echo: true
set.seed(43105)
fit5 <- stan_glm(ldl ~ stat_f, data = class06, refresh = 0)
model_parameters(fit5, ci = 0.90)
```

## LDL means by `statin` group (4/5)

**Bootstrap, without pooling sd estimates**

```{r}
#| echo: true
set.seed(43106)
boot.t.test(ldl ~ stat_f, var.equal = FALSE, R = 2000,
  data = class06, conf.level = 0.90)
```

## LDL means by `statin` group (5/5)

**Bootstrap, with pooling of sd estimates**

```{r}
#| echo: true
set.seed(43107)
boot.t.test(ldl ~ stat_f, var.equal = TRUE, R = 2000,
  data = class06, conf.level = 0.90)
```

## Findings So Far...

Estimating the mean difference in LDL levels at baseline for the "Statin" group minus the "No Statin" group...

Approach | Estimate & 90% Interval
:-------------------------------------: | :------------:
OLS model / Pooled t | 7.17 (0.02, 14.32)
Welch t without pooling sd | 7.17 (0.86, 13.48)
Bayesian fit: `stan_glm()` | 7.26 (0.11, 14.48)
Bootstrap with unpooled sd | 7.16 (1.16, 13.02)
Bootstrap with pooled sd | 7.36 (-0.02, 14.54)

# Standing Break

## LDL by Statin prescription status

To assess how well our data fit the assumptions of the various estimates we have, start with a comparison boxplot.

```{r}
#| echo: true
#| output-location: slide

ggplot(class06, aes(x = ldl, y = stat_f)) +
  geom_violin() +
  geom_boxplot(aes(fill = stat_f), width = 0.3) +
  stat_summary(fun = mean, geom = "point", 
               shape = 18, size = 4, col = "black") +
  guides(fill = "none") +
  labs(y = "Statin prescription", x = "LDL Cholesterol (mg/dl)",
       title = "LDL by Statin prescription",
       subtitle = "Each group is a bit skewed to the right")
```

## LDL by Statin prescription status

How about a pair of Normal Q-Q plots?

```{r}
#| echo: true
#| output-location: slide

p1 <- class06 |>
  filter(stat_f == "Statin") |>
  ggplot(aes(sample = ldl)) +
  geom_qq() + geom_qq_line(col = "red") +
  labs(title = "Statin group", y = "LDL Cholesterol (mg/dl)", 
        x = "Standard Normal")

p2 <- class06 |>
  filter(stat_f == "No") |>
  ggplot(aes(sample = ldl)) +
  geom_qq() + geom_qq_line(col = "red") +
  labs(title = "No Statin group", y = "LDL Cholesterol (mg/dl)", 
        x = "Standard Normal")

p1 + p2 +
  plot_annotation(title = "Normal Q-Q plots of LDL by Statin group")
```


## LDL by Statin prescription status

A look at some numerical summaries...

```{r}
#| echo: true

class06 |> group_by(stat_f) |> reframe(lovedist(ldl)) |> 
  kable(digits = 2) |> kable_styling(font_size = 28)
```

- skew~1~ for "Statin" = (103.43 - 93)/39.51 = 0.26
- skew~1~ for "No" = (96.26 - 91)/31.63 = 0.17

Our rough cutoff for substantial right skew ~ 0.20.

## What can we do about the skew?

- Could we build a confidence interval for medians instead of means?
  - Yes, with the bootstrap, for example see [Section 6.9.2](https://thomaselove.github.io/431-book/06_twogroups.html#bootstrap-ci-for-medians) of my text.

- Could we transform the data in a non-linear way to better match assumptions?
  - Yes, with the help of **Tukey's ladder** of power transformations.

## Tukey's Ladder {.smaller}

Power ($\lambda$) | -2 | -1 | -0.5 | 0 | 0.5 | 1 | 2 | 3
------------: | ---: | ---: | ---: | -----: | ---: | ---: | ---: | ---:
Transformation | $\frac{1}{y^2}$ | $\frac{1}{y}$ | $\frac{1}{\sqrt{y}}$ | log $y$ | $\sqrt{y}$ | $y$ | $y^2$ | $y^3$

- Works to address skew, mainly, rather than symmetry with problems in the tails
- Some of these transformations require all values to be positive. If they aren't, we can add a small amount to each observation until they are.
- Start at no transformation ($\lambda = 1$)
- Try moving in one direction or the other, searching for a result which better matches what we might expect from a Normal distribution
  - If a step (like taking the square root) helps, move in that same direction.
  - If it doesn't help, try moving in the other direction.

## A1c values (end of study)

```{r}
ggplot(class06, aes(x = a1c_end)) + 
  geom_histogram(binwidth = 0.4, fill = "slateblue", col = "white") + 
  stat_function(fun = function(x) 
    dnorm(x, mean = mean(class06$a1c_end, na.rm = TRUE), 
          sd = sd(class06$a1c_end, na.rm = TRUE)) * 
          length(class06$a1c_end) * 0.4,
    geom = "area", alpha = 0.5, 
    fill = "lightblue", col = "blue") +
  labs(title = "A1c (untransformed)", x = "Hemoglobin A1c, %", y = "Count")
```

## Six Power Transformations

```{r}
p1 <- ggplot(class06, aes(x = a1c_end^3)) + 
  geom_histogram(bins = 25, fill = "firebrick", col = "white") + 
  labs(title = "A1c^3 (cubed)", x = "A1c^3", y = "Count")

p2 <- ggplot(class06, aes(x = a1c_end^2)) + 
  geom_histogram(bins = 25, fill = "magenta", col = "white") + 
  labs(title = "A1c^2 (squared)", x = "A1c^2", y = "Count")

p3 <- ggplot(class06, aes(x = a1c_end)) + 
  geom_histogram(bins = 25, fill = "slateblue", col = "white") + 
  labs(title = "A1c (untransformed)", x = "Hemoglobin A1c, %", y = "Count")

p4 <- ggplot(class06, aes(x = sqrt(a1c_end))) + 
  geom_histogram(bins = 25, fill = "skyblue", col = "white") + 
  labs(title = "sqrt(A1c) (square root)", x = "sqrt(A1c)", y = "Count")

p5 <- ggplot(class06, aes(x = log(a1c_end))) + 
  geom_histogram(bins = 25, fill = "navy", col = "white") + 
  labs(title = "log A1c (logarithm)", x = "log A1c", y = "Count")

p6 <- ggplot(class06, aes(x = 1/a1c_end)) + 
  geom_histogram(bins = 25, fill = "thistle", col = "white") + 
  labs(title = "1/A1c (inverse)", x = "1/A1c", y = "Count")

(p1 + p2 + p3) / (p4 + p5 + p6)
```



## Which Transformation to Choose?

Box-Cox approach: can we get a suggested "power" to use when transforming our outcome?

- Specify the model then apply `boxCox` from `car` package.
- Here, all values of our outcome (A1c) are strictly positive. 
  - If not, we'd have to add a constant so that they were.

```{r}
#| echo: true
#| output-location: slide

fit1 <- lm(a1c_end ~ 1, data = class06)
boxCox(fit1)
```

## Back to LDL by Statin prescription

```{r}
ggplot(class06, aes(x = ldl, y = stat_f)) +
  geom_violin() +
  geom_boxplot(aes(fill = stat_f), width = 0.3) +
  stat_summary(fun = mean, geom = "point", 
               shape = 18, size = 4, col = "black") +
  guides(fill = "none") +
  labs(y = "Statin prescription", x = "LDL Cholesterol (mg/dl)",
       title = "LDL by Statin prescription",
       subtitle = "Each group is a bit skewed to the right")
```

## What transformation should we try?

```{r}
#| echo: true

fit2 <- lm(ldl ~ statin, data = class06)
boxCox(fit2)
```

## log(LDL) by Statin?

```{r}
#| echo: true
#| output-location: slide

ggplot(class06, aes(x = log(ldl), y = stat_f)) +
  geom_violin() + 
  geom_boxplot(aes(fill = stat_f), width = 0.3) +
  stat_summary(fun = mean, geom = "point", 
               shape = 18, size = 4, col = "black") +
  labs(y = "Statin prescription", x = "log(LDL)",
       title = "log(LDL) by Statin prescription") +
  guides(fill = "none")
```

## Estimate log(LDL) mean by Statin

```{r}
#| echo: true
#| message: true

fit3 <- lm(log(ldl) ~ stat_f, data = class06)
model_parameters(fit3, ci = 0.90)  ## 90% uncertainty interval
```

- The difference between Statin and No Statin on the logarithmic scale (in mean of log(LDL)) is thus 0.05, with 90% CI (-0.02, 0.12).

## Expectations from our model

- The model coefficients exist in the transformed world, but we can make predictions back on our original scale.

```{r}
#| echo: true

estimate_expectation(fit3, data = "grid", ci = 0.90)
```

- and we exponentiate to get back to our original scale...

## Expectations from our model

- We use `transform = TRUE` to exponentiate back out of the logarithms to our original scale of LDL cholesterol levels.

```{r}
#| echo: true

estimate_expectation(fit3, data = "grid", ci = 0.90, transform = TRUE)
```

## Exponentiation of Expectations

- Back out of logged predictions by exponentiating...

Group | Predicted LDL | 90% uncertainty interval
------: | :----------: | :---------------:
No Statin | exp(4.52) = `r round_half_up(exp(4.52),1)` | (exp(4.46), exp(4.58)) = <br /> (`r round_half_up(exp(4.46),1)`, `r round_half_up(exp(4.58),1)`) 
Statin | exp(4.57) = `r round_half_up(exp(4.57),1)` | (exp(4.54), exp(4.60)) = <br /> (`r round_half_up(exp(4.54),1)`, `r round_half_up(exp(4.60),1)`) 

- These intervals describe uncertainty about the **average predicted LDL across all subjects** in each statin group.

## Fitting Individual Predictions

- Use `estimate_prediction()` to make predictions for an individual subject in each group.

```{r}
#| echo: true

estimate_prediction(fit3, data = "grid", ci = 0.90)
```

- and exponentiate to get back to our original scale...

## Fitting Individual Predictions

- Back out of log transformation with `transform = TRUE`.

```{r}
#| echo: true

estimate_prediction(fit3, data = "grid", ci = 0.90, transform = TRUE)
```

## Fitting Individual Predictions

- Exponentiating the predictions gives...

Group | Predicted LDL | 90% uncertainty interval
------: | :----------: | :---------------:
No Statin | exp(4.52) = `r round_half_up(exp(4.52),1)` | (exp(3.92), exp(5.12)) = <br /> (`r round_half_up(exp(3.92),1)`, `r round_half_up(exp(5.12),1)`) 
Statin | exp(4.57) = `r round_half_up(exp(4.57),1)` | (exp(3.97), exp(5.17)) = <br /> (`r round_half_up(exp(3.97),1)`, `r round_half_up(exp(5.17),1)`) 

- These intervals describe uncertainty about an **individual predicted LDL for a single subject** in each group.

## Using the log transformation

- The log transformation in R (`log` in R is the natural log - the base 10 log is `log10()`) generates coefficients which describe a multiplicative effect.

If we fit a model to predict $y$ using $x$ of the form:

$$
\log y = \beta_0 + \beta_1 x
$$

where $\log y$ is the natural log, then if we exponentiate the slope coefficient, we get the multiplicative factor for a one-unit increase in the predictor.

## The Math

$$
\log(y) = \beta_0 + \beta_1 x \\
\exp(\log(y)) = \exp(\beta_0 + \beta_1 x) \\
y = \exp(\beta_0 + \beta_1 x) \\
y = \exp(\beta_0) \times \exp(\beta_1 x)
$$

This implies that our predictor ($x$) has a multiplicative relationship with our outcome ($y$) instead of the usual additive relationship. Hence, we can express the effect of a one-unit change in $x$ on $y$ as a percentage change.

## Example 1

Suppose we build a regression model with equation $\log y = 6 + 0.23 x$. 

Note that $\exp(0.23) = 1.26$.

- If Harry's value of $x$ is 1 point larger than Sally's, the predicted $y$ value for Harry will be increased by 26% relative to Sally's. (Increased by 26% = multiplied by 1.26.)

## Back to log(LDL) by Statin

```{r}
#| echo: true
fit3 <- lm(log(ldl) ~ stat_f, data = class06)
model_parameters(fit3, ci = 0.90)  ## 90% uncertainty interval
```

- Now, we'll exponentiate these coefficients...

## Exponentiating the Coefficients

```{r}
#| echo: true
fit3 <- lm(log(ldl) ~ stat_f, data = class06)
model_parameters(fit3, ci = 0.90, exponentiate = TRUE)  
```

Subjects *with* a statin prescription have 5% higher LDL, on average, than do subjects *without*. 

- The ratio has point estimate 1.05 and 90% interval estimate (0.98, 1.13).

## Independent Samples Strategy {.smaller}

Suppose we want to estimate an uncertainty interval for the difference in means across two groups. To begin, plot the data from each sample.

- If the sample data in each group are well described as "Normal", then use the OLS / pooled t procedure to obtain a confidence interval, or a Bayesian model to obtain a credible interval.
    - If the sample sizes are not the same, and the sample variances aren't close to each other, consider using a Welch t procedure.
- If either sample's data are "symmetric but with outliers", then consider using the bootstrap to obtain a confidence interval, although an OLS or Bayesian result may also be reasonable.
- If either sample's data are best described as "substantially skewed", then consider whether you actually want to summarize with the difference in means, and consider whether a transformation might be helpful.

# Another worked example (for self-study)

## Boxplot: Difference in A1c by Sex

```{r}
#| echo: true

ggplot(class06, aes(x = a1c_diff, y = sex)) +
  geom_violin() + geom_boxplot(aes(fill = sex), width = 0.3) +
  labs(x = "Change in A1c", y = "Sex",
       title = "Difference in A1c (end - base) by Sex") +
  guides(fill = "none")
```

## Numerical Summary (A1c by Sex)

```{r}
#| echo: true
class06 |> group_by(sex) |> 
  reframe(lovedist(a1c_diff)) |> 
  kable(digits = 3) |> kable_styling(font_size = 28)
```

- Difference in sample means is 0.503 - 0.227 = 0.276
- Do we have a balanced design?
- Are we comfortable pooling standard deviations?
- Are we comfortable assuming Normality for each sex?

## Might a transformation help here?

Some of the A1c changes are negative (minimum = -5.9)

```{r}
#| echo: true

class06 <- class06 |> mutate(a1c_diff_p6 = a1c_diff + 6)
fit4 <- lm(a1c_diff_p6 ~ sex, data = class06)
boxCox(fit4)
```

## Does using a square root help?

```{r}
#| echo: true
#| output-location: slide

p1 <- ggplot(class06, aes(x = a1c_diff_p6, y = sex)) +
  geom_violin() + geom_boxplot(aes(fill = sex), width = 0.3) +
  labs(x = "Change in A1c + 6", y = "Sex",
       title = "A1c difference (+ 6) by Sex") +
  guides(fill = "none")

p2 <- ggplot(class06, aes(x = sqrt(a1c_diff_p6), y = sex)) +
  geom_violin() + geom_boxplot(aes(fill = sex), width = 0.3) +
  labs(x = "Square Root of (Change in A1c + 6)", y = "Sex",
       title = "Square Root of (A1c difference + 6) by Sex") +
  guides(fill = "none")

p1 / p2
```

## Addressing outliers?

### OLS model

```{r}
#| echo: true

fit5 <- lm(a1c_diff ~ sex, data = class06)

model_parameters(fit5, ci = 0.90)
```

- Male subjects had A1c changes that were 0.28 larger than females, on average, with 90% uncertainty interval (0, 0.56) according to our OLS model (pooled t test.)

## Addressing outliers?

### Bayesian model

```{r}
#| echo: true
set.seed(20240912)
fit6 <- stan_glm(a1c_diff ~ sex, data = class06, refresh = 0)

model_parameters(fit6, ci = 0.90)
```

- Male subjects had A1c changes that were 0.28 larger than females, on average, with 90% uncertainty interval (0, 0.57) according to our Bayesian model.

## Addressing outliers?

### Bootstrap with `boot.t.test`

```{r}
#| echo: true
set.seed(20240912)
boot.t.test(a1c_diff ~ sex, data = class06, conf.level = 0.90)
```

## Repeating: Paired Samples Strategy {.smaller}

Suppose we want to estimate an uncertainty interval for the mean of a set of paired differences.

- Calculate the paired differences, then plot them.
- If the sample data are well described as "Normal", then use the OLS / paired t procedure to obtain a confidence interval, or a Bayesian model to obtain a credible interval.
- If the sample data are best described as "symmetric but with outliers", then use the bootstrap to obtain a confidence interval, although an OLS or Bayesian result may be quite similar.
- If the sample data are best described as "substantially skewed", then consider whether you actually want to summarize with the mean, and consider whether a transformation might be helpful.

## Independent Samples Strategy {.smaller}

Suppose we want to estimate an uncertainty interval for the difference in means across two groups. To begin, plot the data from each sample.

- If the sample data in each group are well described as "Normal", then use the OLS / pooled t procedure to obtain a confidence interval, or a Bayesian model to obtain a credible interval.
    - If the sample sizes are not the same, and the sample variances aren't close to each other, consider using a Welch t procedure.
- If either sample's data are "symmetric but with outliers", then consider using the bootstrap to obtain a confidence interval, although an OLS or Bayesian result may also be reasonable.
- If either sample's data are best described as "substantially skewed", then consider whether you actually want to summarize with the difference in means, and consider whether a transformation might be helpful.

## Session Information

```{r}
#| echo: true
xfun::session_info()
```

