---
title: "431 Class 10"
author: Thomas E. Love, Ph.D.
date: "2026-09-24"
format:
  revealjs: 
    theme: dracula
    embed-resources: true
    self-contained: true
    slide-number: true
    footnotes-hover: true
    preview-links: auto
    date-format: iso
    footer: "431 Class 10 | 2026-09-24 | <https://thomaselove.github.io/431-2026/>"
---


## Today's Agenda

- Checking the fit of a linear model to assumptions
  - linearity
  - constant variance (homoscedasticity)
  - Normality of residuals
    - assessing influence and leverage
  - posterior predictive checks
- The impact of a transformation
- Model performance measures besides $R^2$ and $\sigma$

## Today's Setup

```{r}
#| echo: true
#| message: false

knitr::opts_chunk$set(comment = NA)

source("c10/data/Love-431.R") # for the lovedist() function

library(janitor)
library(naniar)     ## dealing with missing data
library(car)        ## for the Box-Cox plot 
library(ggExtra)    ## NEW! add margins to scatterplot
library(patchwork)
library(rstanarm)
library(easystats)
library(tidyverse)

theme_set(theme_light()) ## new theme for ggplots!
```

## Data Ingest and Cleaning

[EUROFAMCARE](https://www.uke.de/extern/eurofamcare/) data introduced in Class 09.

- Our data describe 902 primary caregivers providing at least four hours of personal care or support per week to an elder aged 65 years or older.

```{r}
#| echo: true

eurocare <- read_csv("c10/data/eurocare.csv", show_col_types = FALSE) |>
  janitor::clean_names() |>
  mutate(subject = as.character(subject)) |>
  select(subject, e_age, barthel, neg_impact, pos_impact)

dim(eurocare)
```

## `eurocare` variables

Variable | Description
-------- | -----------------------------------------------------
`subject` | Identifying code
`e_age` | Age of elder (years)
`barthel` | [Barthel index](https://en.wikipedia.org/wiki/Barthel_scale) for elder's Activities of Daily Living (0 to 100 = complete independence)
`neg_impact` | COPE Index Negative Impact subscale
`pos_impact` | COPE Index Positive Impact subscale

## Missingness Reports

```{r}
#| echo: true

miss_case_table(eurocare) # missingness by subject

miss_var_summary(eurocare) # missingness by variable
```

## Filter to complete cases

- Drop missing values across all variables in `eurocare`.

```{r}
#| echo: true
ecare <- eurocare |> drop_na()
ecare
```

## Automated codebook

```{r}
#| echo: true

data_codebook(ecare |> select(-subject))
```

## Negative vs. Positive Impact

```{r}
#| echo: true
#| output-location: slide

plot1 <- ggplot(ecare, aes(x = pos_impact, y = neg_impact)) +
  geom_jitter() +
  geom_smooth(method = "lm", col = "red", formula = y ~ x, se = TRUE) +
  geom_smooth(method = "loess", col = "blue", formula = y ~ x, se = FALSE) +
  labs(x = "Positive Impact", y = "Negative Impact")

plot1
```

## Add margins to scatterplot

```{r}
#| echo: true
ggMarginal(plot1, type = "histogram")
```



## Model we fit last time

```{r}
#| echo: true

fit3 <- lm(neg_impact ~ pos_impact, data = ecare)
model_parameters(fit3, ci = 0.97, pretty_names = FALSE)
model_performance(fit3)
```

## Checking OLS Model Fit

Main assumptions of any simple linear regression are:

1. **linearity**: we assume that the outcome is linearly related to our predictor 
2. **constant variance** (homoscedasticity): we assume that the variation of our outcome is about the same regardless of the value of our predictor
3. **normal distribution**: we assume that the errors around the regression model at any specified values of the x-variables follow an approximately Normal distribution.

## Fitting the diagnostic plots

To check these assumptions, consider the following plots.

```{r}
#| echo: true
#| message: true

fit3_diagnostic_plots <- 
  plot(check_model(fit3, panel = FALSE, detrend = FALSE))
```

- I don't worry about confidence bands in these plots.

## Checking Linearity

```{r}
#| echo: true

fit3_diagnostic_plots[[2]]
```

## Checking Equal Variances

```{r}
#| echo: true

fit3_diagnostic_plots[[3]]
```

## Checking for Influential Points

```{r}
#| echo: true

fit3_diagnostic_plots[[4]]
```

## Checking Normality

```{r}
#| echo: true

fit3_diagnostic_plots[[5]]
```

## Alternative: Normal Q-Q

```{r}
#| echo: true

plot(fit3, which = 2)
```

## Posterior Predictive Checks

```{r}
#| echo: true

fit3_diagnostic_plots[[1]]
```

## Box-Cox suggestion?

```{r}
#| echo: true
boxCox(fit3)
```

## Transformation to Scatterplot

```{r}
#| echo: true

plot2 <- ggplot(ecare, aes(x = pos_impact, y = (1/sqrt(neg_impact)))) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, col = "red", formula = y ~ x)

plot2
```

## Add margins to scatterplot

```{r}
#| echo: true
ggMarginal(plot2, fill = "purple", type = "boxplot")
```

## What would a transformed model look like?

```{r}
#| echo: true

ecare <- ecare |> mutate(neg_tr = 1/sqrt(neg_impact))

fit4 <- lm(neg_tr ~ pos_impact, data = ecare)

fit4_diagnostic_plots <- 
  plot(check_model(fit4, panel = FALSE, detrend = FALSE))
```

## Posterior Predictive Check?

```{r}
#| echo: true

fit4_diagnostic_plots[[1]]
```

## Normality of Residuals?

```{r}
#| echo: true

fit4_diagnostic_plots[[5]]
```


## Interpreting the transformed model

```{r}
#| echo: true

model_parameters(fit4, ci = 0.97)

model_performance(fit4)
```

# Standing Break

## A new question

How well can we predict an elder's `barthel` index (0 = complete dependence for activities daily living to 100 = complete independence) on the basis of their age (stored in `e_age`)?

```{r}
#| echo: true
ecare |> reframe(lovedist(barthel)) 
ecare |> reframe(lovedist(e_age))
```

## Histograms?

```{r}
#| echo: true
#| output-location: slide

p1 <- ggplot(ecare, aes(x = barthel)) +
       geom_histogram(binwidth = 1, fill = "blue", col = "white") + 
       labs(title = "Barthel Index")

p2 <- ggplot(ecare, aes(x = e_age)) +
       geom_histogram(binwidth = 1, fill = "forestgreen", col = "white") + 
       labs(title = "Age of elder")

p1 + p2
```

## Scatterplot of 847 elders

We'll look at `barthel` index (on the y axis) vs. `e_age` (on x). We'll plot each point with `alpha = 0.3` so that if we have multiple overlapping points, the plot will look darker there.

```{r}
#| echo: true
#| output-location: slide

ggplot(ecare, aes(x = e_age, y = barthel)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", 
              formula = y ~ x, se = FALSE, col = "blue") +
  geom_smooth(method = "lm", 
              formula = y ~ x, se = TRUE, col = "red") +
  labs(title = "EUROFAMCARE data for 847 subjects",
       x = "Age (years)", y = "Barthel Index for ADL",
       caption = "Linear model in red, loess smooth in blue")
```

## Jittered Scatterplot?

Let's add a little random noise to each point on each scale to deal with the problem of overlapping points. We use `geom_jitter()` here rather than `geom_point()`.

```{r}
#| echo: true
#| output-location: slide

ggplot(ecare, aes(x = e_age, y = barthel)) +
  geom_jitter() +
  geom_smooth(method = "loess", 
              formula = y ~ x, se = FALSE, col = "blue") +
  geom_smooth(method = "lm", 
              formula = y ~ x, se = TRUE, col = "red") +
  labs(title = "Jittered EUROFAMCARE data for 847 subjects", 
       x = "Age (years)", y = "Barthel Index for ADL",
       caption = "Linear model in red, loess smooth in blue")
```

## Pearson Correlation Coefficient

- Pearson $r$ ranges from -1 (perfect inverse relationship: as x goes up, y goes down) to +1 (perfect direct relationship: as x goes up, y goes up.)
- Pearson $r$ measures only a *linear* association.

```{r}
#| echo: true

ecare |> select(barthel, e_age) |> correlation()
```

## OLS: Predict `barthel` with `e_age`

```{r}
#| echo: true

fit5 <- lm(barthel ~ e_age, data = ecare)
model_parameters(fit5, ci = 0.89)
```

- Note that our variable `e_age` has been "prettified" here to `e age` which might be confusing...

```{r}
#| echo: true
model_parameters(fit5, ci = 0.89, pretty_names = FALSE)
```

What does this output tell us?

## Interpreting the Intercept

```{r}
#| echo: true
model_parameters(fit5, ci = 0.89, pretty_names = FALSE)
```

- Our fit5 model estimates the `barthel` value of to be 109.01 if `e_age` = 0, but we have no values of `e_age` anywhere near 0 in our data, nor is such a value plausible clinically for either `e_age` (our values range from 65 to 103) or `barthel` (which must be between 0 and 100), so the intercept doesn’t tell us much here.

## Point Estimate of the Slope

```{r}
#| echo: true
model_parameters(fit5, ci = 0.89, pretty_names = FALSE)
```

- Suppose Harry is one year older than Sally. Our `fit5` model predicts the `barthel` index for Harry to be 0.55 points *less* than that of Sally, on average.
- Of course, this assumes Harry and Sally are similar to the subjects in `e_care` and that our linear model `fit5` is correct.

## Confidence Interval for Slope

```{r}
#| echo: true
model_parameters(fit5, ci = 0.89, pretty_names = FALSE)
```

- When we generalize beyond study participants to the population they were selected at random from, then our data are compatible (at the 89% confidence level) with population slopes between -0.75 and -0.36, depending on the assumptions of our linear model `fit5` being correct.

## Performance of our OLS model?

```{r}
#| echo: true

model_performance(fit5)
```

- Key summaries for an OLS model with one predictor, like `fit5`, are $R^2$ and Sigma (which is similar to RMSE.)
- $R^2$ tells us model `fit5` accounts for 2.4% of the variation in `barthel` that we observe in our `ecare` data.

## Using the estimated `Sigma` ($\sigma$)

```{r}
#| echo: true

model_performance(fit5)
```

- Our model `fit5` assumes that our errors (residuals) come from a Normal distribution with mean 0 and standard deviation Sigma ($\sigma$) = 28.8. 
- Thus, about 68% of our predictions should be within $\pm$ 28.8 points of the correct `barthel` index, and 95% of our predictions should be within $\pm 2 \times 28.8$, or 57.6 points on the `barthel` score. Is this good news?

## Performance Measures (1/5)

- `AIC`: Akaike’s Information Criterion
- `AICc`: Second-order (or small sample) AIC with a correction for small sample sizes
- `BIC`: Bayesian Information Criterion

AIC, AICc and BIC are used when comparing one or more models for the same outcome. When comparing models fit using maximum likelihood (like OLS linear models), the smaller the AIC or BIC, the better the fit.

## Performance Measures (2/5)

`R2`: r-squared value = 0.024

- The R-squared ($R^2$) measure for an OLS fit describes how much of the variation in our outcome can be explained using our model (and its predictors.) $R^2$ falls between 0 and 1, and the closer it is to 1, the better the model fits our data. 
- In a simple (one-predictor) OLS model like this, the value is also the square of the Pearson correlation coefficient, $r$. 
- We called this $R^2$ "eta-squared" ($\eta^2$) in ANOVA.

## Performance Measures (3/5)

`R2 (adj.)`: adjusted r-squared value = 0.023

- Adjusted R-squared is an index (so it’s not a proportion of anything) for comparing different models (different predictor sets) for the same outcome.
- The idea is to reduce the temptation to overfit the data, by penalizing the $R^2$ value a little for each predictor. 
- Adjusted $R^2$ is usually between 0 and 1, but can be negative.
- The adjusted $R^2$ measure can never be larger than $R^2$.

## Performance Measures (4/5)

`RMSE` = 28.742

- The RMSE is the square root of the variance of the residuals and summarizes the difference between the observed data and the model’s predicted values.
- It can be interpreted as the standard deviation of the unexplained variance, and has the same units as the outcome. 
- When comparing models using the same data for the same outcome (but, for instance, with different predictor sets), lower RMSE values indicate better model fit. 

## Performance Measures (5/5)

`Sigma` = 28.776

- Linear models assume that their residuals are drawn from a Normal distribution with mean 0 and standard deviation equal to sigma ($\sigma$). 
- This indicates that the predicted outcome will be within $\pm 1 \sigma$ units of the observed outcome for approximately 68% of the data points, for example. 

## Checking OLS Model Fit

Main assumptions of any simple linear regression are:

1. **linearity**: we assume that the outcome is linearly related to our predictor 
2. **constant variance** (homoscedasticity): we assume that the variation of our outcome is about the same regardless of the value of our predictor
3. **normal distribution**: we assume that the errors around the regression model at any specified values of the x-variables follow an approximately Normal distribution.

## Fitting the diagnostic plots

To check these assumptions, consider the following plots.

```{r}
#| echo: true
#| message: true

fit5_diagnostic_plots <- 
  plot(check_model(fit5, panel = FALSE, detrend = FALSE))
```

- I don't worry about confidence bands in these plots.

## Checking Linearity

```{r}
#| echo: true

fit5_diagnostic_plots[[2]]
```

## Checking Equal Variances

```{r}
#| echo: true

fit5_diagnostic_plots[[3]]
```

## Checking for Influential Points

```{r}
#| echo: true

fit5_diagnostic_plots[[4]]
```

## Which points are listed in the plot?

None are anywhere near the contours, as it turns out.

```{r}
#| echo: true
ecare |> slice(c(4, 38, 160, 470)) |> select(subject, barthel, e_age)
```

## Checking Normality

```{r}
#| echo: true

fit5_diagnostic_plots[[5]]
```

## Alternative: Normal Q-Q

```{r}
#| echo: true

plot(fit5, which = 2)
```

## Which points are low outliers?

```{r}
#| echo: true
ecare |> slice(c(4, 47, 160)) |> select(subject, barthel, e_age)
```

## Posterior Predictive Checks

```{r}
#| echo: true

fit5_diagnostic_plots[[1]]
```

## Box-Cox suggestion?

To use Box-Cox, all values of our outcome `barthel` would have to be strictly positive. 

```{r}
#| echo: true

ecare |> reframe(lovedist(barthel)) 
```

```{r}
#| echo: true
#| output-location: slide
fit5_adj <- lm((barthel+1) ~ e_age, data = ecare)

boxCox(fit5_adj)
```

## Next Time

What's wrong with statistical significance?

## Session Information

```{r}
#| echo: true

xfun::session_info()
```

